---
title: "Detection methodology"
subtitle: "The moving-echo truck detector — architecture, mathematics, inputs and outputs"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    number_sections: true
    theme: flatly
    df_print: kable
    fig_width: 9
    fig_height: 5
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 150, fig.retina = 2)
suppressPackageStartupMessages({library(readr); library(dplyr); library(ggplot2); library(tibble)})
theme_set(theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(colour = "grey35", size = 10),
        plot.caption = element_text(colour = "grey45", size = 9, hjust = 0),
        legend.position = "top", legend.title = element_blank()))
```

# The model

The detector is a **Keypoint R-CNN** — a two-stage convolutional object detector with an additional head that
localises a fixed number of landmark points on every detected object. It is implemented with the torchvision
reference model `keypointrcnn_resnet50_fpn`, using a ResNet-50 backbone with a Feature Pyramid Network, and
comprises 59.1 million trainable parameters.

Two trained instances are in current use. `kprcnn-adamiak-v2` was trained in July 2026 on nine scenes and serves
as the project baseline. `kprcnn-warmup-v1` was trained later on twenty scenes with learning-rate warmup,
translation-jitter augmentation and multi-vehicle chip targets. Both share the architecture described here and
differ only in training data and optimisation schedule.

The model is configured for two classes — a single foreground class named `moving_echo` plus the implicit
background class — and three keypoints per detection. It is trained and run entirely on CPU, because the
detection operations it depends on are numerically unstable on the Apple Metal backend available on the
development machine.

# What the model is looking for

PlanetScope SuperDove satellites do not expose their spectral bands simultaneously. Each band is read from a
different row of the sensor's focal plane, so the satellite advances slightly between one band's exposure and the
next. For stationary ground this is invisible, because the bands register on top of one another. For a moving
vehicle it is not: the vehicle occupies a different ground position in each exposure, so rather than appearing as
a single object it smears into a short colour-separated streak — a blue blob, then a red blob, then a green blob,
strung out along the direction of travel. The readout order is blue, then red, then green, which makes the streak
directional and therefore informative about heading.

This matters because of scale. At three metres per pixel a passenger car covers roughly one pixel and a truck two
to six. Detecting an object that small directly is close to hopeless. The streak it leaves, however, spans a far
larger footprint — a median of about 34 metres end to end in this dataset, roughly ten times the length of the
vehicle producing it — and has a tri-colour structure that natural terrain rarely reproduces. The detector
therefore does not search for trucks. It searches for the three-point colour signature a moving truck leaves
behind, and infers the truck from it.

```{r fig-streak}
det <- read_csv("data/detections_warmup-v1.csv", show_col_types = FALSE,
                col_select = c(labelled, is_window_top1, matched_bool, streak_len_m)) |>
  filter(labelled == 1, is_window_top1 == 1, matched_bool == 1)
ggplot(det, aes(streak_len_m)) +
  geom_histogram(bins = 45, fill = "#2a78d6", alpha = 0.85) +
  geom_vline(xintercept = median(det$streak_len_m), linetype = "dashed", colour = "grey25") +
  annotate("text", x = median(det$streak_len_m) + 3, y = Inf, vjust = 1.6, hjust = 0,
           label = sprintf("median %.0f m", median(det$streak_len_m)), size = 3.4, colour = "grey25") +
  scale_x_continuous(limits = c(0, 90)) +
  labs(title = "Echo length, blue keypoint to green keypoint",
       subtitle = "Detections that matched a hand label; a pixel is three metres.",
       x = "streak length (metres)", y = "detections")
```

The streak's length is proportional to the vehicle's speed, since a faster truck travels further between
exposures. Converting that into an absolute velocity requires the inter-band time interval for each scene, which
has not yet been obtained, so velocity recovery remains unimplemented.

# Inputs

## Imagery

The imagery is PlanetScope `PSScene` data in the `analytic_8b_sr_udm2` bundle, captured by the PSB.SD SuperDove
instrument. Each scene is an eight-band unsigned 16-bit GeoTIFF of Surface Reflectance at three metres per pixel,
delivered in a UTM projection — EPSG:32610 for western Washington and EPSG:32611 for eastern — with zero used as
the nodata value outside the clipped corridor.

The imagery is never reprojected, warped or resampled at any point in the pipeline. This is the single most
important input constraint, and it follows directly from the nature of the signal: the thing being measured is a
sub-pixel positional offset between bands, and any resampling interpolates neighbouring pixels together and
smears that offset away. Scenes are consequently kept on their native grid in their native zone. Where a
coordinate transform is unavoidable — because the annotation file and the raster disagree about projection — the
annotation points are moved rather than the raster, since transforming point coordinates is exact arithmetic on a
handful of numbers rather than a lossy interpolation over millions of pixels.

Only three of the eight bands reach the model. Band 6 becomes the red channel, band 4 the green, and band 2 the
blue, which reconstructs a true-colour image and, not coincidentally, captures the three exposures that form the
echo.

```{r bands}
tibble(
  band = 1:8,
  name = c("Coastal Blue", "Blue", "Green I", "Green", "Yellow", "Red", "Red Edge", "NIR"),
  role = c("unused", "blue channel — first exposure", "unused", "green channel — third exposure",
           "unused", "red channel — second exposure", "unused", "unused")
)
```

The five unused bands carry no part of the colour-separation signature, though whether red-edge or near-infrared
contain independent information about the echo has never been tested.

## Annotations

Labels are stored in a GeoPackage as points in real-world coordinates rather than pixel coordinates, one row per
keypoint. Each row carries a point geometry, an integer `vehicle_id` grouping the three points belonging to one
vehicle, an integer `sequence` recording which exposure the point belongs to — one for blue, two for red, three
for green — and a `scene` string naming the GeoTIFF the point belongs to. A vehicle is therefore exactly three
rows sharing a `vehicle_id`, and any vehicle missing one of the three sequences is treated as incomplete and
dropped during export.

The labelling policy is deliberately conservative: only clear, well-formed echoes are marked, while blurry,
malformed, over-bright and very small echoes are deliberately skipped. This decision propagates into every
downstream metric, because the imagery contains real vehicles that are knowingly unlabelled. A detection landing
on one of those is scored as a false positive even though the model was correct, which means reported precision
is a lower bound rather than an estimate.

# From reflectance to tensors

Converting a scene into something the network can consume happens in two stages: a radiometric stretch that maps
16-bit reflectance into an 8-bit image, and a geometric join that converts annotation coordinates into pixel
positions.

The stretch operates per scene and per band. For each of the three selected bands the second and ninety-eighth
percentiles of the non-zero pixels are computed, and reflectance values are mapped linearly from that range onto
zero to 255, with anything outside clipped to the endpoints. Formally, for a band $b$ with percentile bounds
$p_{2}$ and $p_{98}$, a pixel value $v$ becomes

$$v' = 255 \cdot \mathrm{clip}\!\left(\frac{v - p_{2}}{p_{98} - p_{2}},\; 0,\; 1\right)$$

with nodata pixels forced to zero rather than stretched. Because the percentiles are recomputed for every scene,
each scene is contrast-normalised against its own distribution. This makes scenes captured under different
illumination or atmospheric conditions comparable to one another, at the cost that identical physical reflectance
maps to different pixel values in different scenes. The same stretch code runs during training and during
inference, so there is no mismatch between the two.

The join converts map coordinates into pixel coordinates. Each annotation is first matched to its scene by the
`scene` text field — never by spatial extent, because several scenes cover the same ground on different dates and
a geometric join would assign one date's labels to another date's imagery. The points are then reprojected into
the raster's own CRS if the two differ, and finally the raster's inverse affine transform converts each map
coordinate into a fractional pixel column and row. Keypoints are carried as floating-point values throughout;
rounding them to integers would introduce a systematic sub-pixel error in precisely the quantity the model is
trying to predict.

Finally, the 8-bit image is divided by 255 to produce floats in the unit interval, and torchvision's internal
transform applies ImageNet channel normalisation — subtracting means of 0.485, 0.456 and 0.406 and dividing by
standard deviations of 0.229, 0.224 and 0.225 — before the tensor reaches the backbone. This second
normalisation exists because the backbone's pretrained weights were learned on data normalised that way.

# Chip extraction

The network trains on small crops rather than whole scenes. For every labelled vehicle the exporter cuts a
96 × 96 pixel window centred on the mean of that vehicle's three keypoints, writes it as a PNG, and records the
keypoints relative to the window's origin along with a bounding box derived from the keypoint extent plus three
pixels of padding on each side. The current corpus of 789 labelled vehicles therefore produces 789 chips.

The window is 96 pixels but the model consumes 64. The extra sixteen pixels on each side exist so the training
loop can crop a 64-pixel window at a randomly chosen offset on every epoch, placing the vehicle at a different
position within the chip each time it is seen. Without that margin every training example would have its vehicle
at dead centre, and the network would learn that echoes appear in the middle of the frame — a rule that fails
immediately at inference, where a sliding window presents vehicles at arbitrary offsets.

Any other labelled vehicle falling wholly inside the same 96-pixel window is also annotated, which adds 497
further targets on top of the 789 centred ones. Without this, a second truck sharing a chip with the first would
be unlabelled, and the network would be explicitly taught that a real echo is background.

One structural consequence deserves emphasis. Because every chip is centred on a vehicle, the background the
network learns is always road surface and roadside terrain within roughly a hundred metres of a truck. It never
sees field interior, rooftops or bare soil presented as labelled background, so its notion of "not a truck" is
narrower than the terrain it must actually operate over.

# Architecture

## Backbone and feature pyramid

The first stage is a ResNet-50 convolutional backbone wrapped in a Feature Pyramid Network. ResNet-50 processes
the image through successive residual blocks, halving spatial resolution and doubling channel depth at each
stage, so that later layers encode more abstract structure at coarser resolution. The Feature Pyramid Network
then takes the outputs of several stages and combines them top-down, upsampling coarse semantic features and
adding them to finer ones, producing five feature maps at different scales that all carry high-level semantic
information. Small objects can therefore be detected on high-resolution maps without sacrificing the semantic
depth that only deeper layers provide.

The backbone enters training with weights pretrained on COCO. The detection heads do not, because their output
dimensions depend on the number of classes and keypoints, which differ from COCO's 91 classes and 17 human body
keypoints. Both heads are replaced and randomly initialised.

## Region proposal network

The second stage proposes candidate object locations. At every spatial position of every pyramid level the
network places a set of reference boxes called anchors, and a small convolutional head predicts, for each anchor,
an objectness logit and four numbers describing how to adjust the anchor into a better-fitting box.

Anchors are the most consequential setting for this problem. The torchvision defaults span 32 to 512 pixels,
which is far too large for a two-to-six-pixel echo, so a custom set of five sizes — 4, 8, 16, 32 and 48 pixels,
one per pyramid level — is used, crossed with five aspect ratios of 0.25, 0.5, 0.75, 1.0 and 1.25. That gives 25
anchors per spatial location.

A subtlety affects how those numbers should be read. Chips are 64 pixels but torchvision resizes its input so the
shorter side is 192 pixels, a threefold upscale. Anchors therefore operate in resized space, not native chip
space: a nominally four-pixel anchor corresponds to about 1.3 native pixels, while a native four-to-eight-pixel
echo becomes twelve to twenty-four pixels after resizing and is matched by the middle anchors rather than the
smallest ones. The resize is already doing much of the scale-matching work that small anchors were introduced to
provide, and any future anchor tuning must reason in resized space to avoid compounding the two effects.

During training each anchor is assigned a label by its intersection-over-union with the ground-truth boxes. An
anchor overlapping a target by more than 0.7 is positive, one overlapping by less than 0.3 is negative, and
anything between is ignored. From these, 256 anchors are sampled per image with a target of half positive, and
the objectness head is trained with binary cross-entropy on that sample. Box regression uses the standard
parameterisation, in which the network predicts offsets relative to the anchor rather than absolute coordinates:

$$t_x = \frac{x - x_a}{w_a}, \qquad t_y = \frac{y - y_a}{h_a}, \qquad t_w = \log\frac{w}{w_a}, \qquad t_h = \log\frac{h}{h_a}$$

where $(x, y, w, h)$ is the target box and the subscript $a$ denotes the anchor. Predicting offsets normalised by
anchor size makes the regression scale-invariant, and the logarithm on width and height keeps the target
symmetric between growing and shrinking. The regression is trained with a smooth $L_1$ loss, which behaves
quadratically near zero and linearly further out, making it less sensitive to outliers than squared error. Only
positive anchors contribute. At inference the proposals are ranked by objectness, the top thousand are kept, and
non-maximum suppression at an IoU of 0.7 removes duplicates.

## Box head

Surviving proposals pass to the box head. Because proposals have arbitrary sizes and neural networks need fixed
input dimensions, features are extracted with RoIAlign, which samples a fixed 7 × 7 grid from each proposal's
region of the feature map using bilinear interpolation. RoIAlign avoids the quantisation that its predecessor
RoIPool introduced by rounding region boundaries to integer positions — a rounding error of half a pixel, which
is negligible for large objects but material when the object is only a few pixels across.

The 7 × 7 × 256 output is flattened into a 12,544-element vector and passed through two fully connected layers of
1024 units each, from which two linear heads emerge: one producing class logits and one producing box refinements
for each class. Classification is trained with softmax cross-entropy over the two classes, and refinement with
smooth $L_1$ on the same offset parameterisation used by the proposal network, though with the deltas scaled by
weights of 10, 10, 5, 5 so translation and size errors contribute comparably to the gradient. Proposals are
assigned to targets at an IoU threshold of 0.5, and 512 are sampled per image with a quarter positive.

## Keypoint head

The keypoint head is what distinguishes this model from an ordinary detector, and it is where the moving-echo
structure is actually recovered.

Features are pooled from each detection at 14 × 14 resolution and passed through eight convolutional layers of
512 channels, then a transposed convolution upsamples the result and a final interpolation brings it to
**56 × 56**. The output is three such maps per detection — one for the blue keypoint, one for red, one for green.

The important design decision is that keypoint localisation is posed as **classification, not regression**. Each
56 × 56 map is flattened into 3,136 positions, and the network is trained to predict which single position
contains the keypoint, using softmax cross-entropy against a one-hot target. Writing $\ell_{k,i}$ for the logit
of keypoint $k$ at position $i$ and $i^{*}_{k}$ for the true position, the loss is

$$\mathcal{L}_{\text{kp}} = -\frac{1}{K}\sum_{k=1}^{K} \log \frac{\exp\!\left(\ell_{k,\,i^{*}_{k}}\right)}{\sum_{j=1}^{3136} \exp\!\left(\ell_{k,\,j}\right)}$$

Treating position as a categorical variable rather than a continuous one lets the network express uncertainty
across the whole spatial field and produces sharper localisation than direct coordinate regression, which tends
to average over ambiguity. At inference the predicted keypoint is the arg-max position of its heatmap, and the
logit value at that peak is returned as the keypoint's confidence.

That confidence is worth dwelling on. It is the raw pre-softmax logit at the peak, so it is unbounded rather than
a probability — values in this dataset range from about −4 to +8 — and only its ordering is meaningful. It
measures how sharply peaked the heatmap is, which is a different question from the one the box classifier answers.
A patch of bright field texture can look sufficiently object-like to score well on the box head while producing
three flat, diffuse keypoint maps, because there is no crisp blue-red-green triple to lock onto. Empirically the
keypoint confidence separates correct from spurious detections better than the box score does, with an ROC area
of 0.891 against 0.771, despite the fact that the deployed inference code reads only the box score and discards
the keypoint confidences entirely.

## The composite loss

Training minimises the unweighted sum of five terms: the proposal network's objectness and box-regression losses,
the box head's classification and refinement losses, and the keypoint head's cross-entropy. No relative weighting
is applied, so the gradient balance between localisation and classification is whatever the default magnitudes
happen to produce. Tuning those weights is an untried avenue.

# Training procedure

Optimisation uses Adam at a base learning rate of $10^{-3}$, with gradients clipped to a global norm of 1.5 to
prevent occasional large updates from destabilising the fresh heads. Batches contain four chips, and each vehicle
is drawn three times per epoch with independently sampled augmentation. Runs last twelve to sixteen epochs.

The learning rate is not constant. For the first 300 iterations it ramps linearly from $10^{-5}$ to $10^{-3}$.
This warmup exists because the detection heads begin from random weights, and their initial gradients at full
rate are large enough to destabilise the model — in one run without warmup the loss climbed from 9 to 528 within
two iterations, which forced a permanent tenfold reduction in learning rate and left that model undertrained.
After warmup, `ReduceLROnPlateau` monitors validation loss and multiplies the rate by 0.3 whenever two
consecutive epochs fail to improve, down to a floor of $10^{-5}$.

Augmentation is applied to the full 96-pixel chip before the 64-pixel crop is taken, so that rotation draws in
genuine neighbouring imagery rather than mirrored padding. Horizontal and vertical flips are each applied with
probability one half, rotation is uniform over the full circle, a mild perspective warp displaces the corners by
up to five pixels, and brightness is scaled multiplicatively between 0.8 and 1.2. Every geometric transform is
applied to the keypoint coordinates as well as to the pixels. Translation is handled separately, by the random
crop offset rather than by warping, so it introduces no interpolation at all — the vehicle simply appears at a
different place in a window of real pixels.

Validation is a random subset of chips, the larger of eight or twelve percent, drawn from the training scenes. It
drives both the learning-rate schedule and the selection of which epoch to keep. This is a known weakness:
because validation shares scenes with training, it measures fit to already-familiar backgrounds rather than
generalisation, so best-epoch selection is biased toward models that have memorised training-scene context.
Splitting validation by scene rather than by chip is a recommended change that has not yet been made.

# Inference

Training operates on 64-pixel chips centred on vehicles. Deployment must find vehicles in a scene of arbitrary
size with no prior knowledge of where they are, and that gap is bridged with a sliding window.

The scene is rendered through the same band selection and percentile stretch used in training. Pixels where all
three channels are zero are marked as nodata. A grid of 64 × 64 windows is then placed across the scene at a
stride of 40 pixels, giving 24 pixels of overlap, and a window is evaluated only if at least fifteen percent of
it contains valid imagery. Surviving windows are batched twelve at a time through the network.

The overlap is deliberate. A vehicle is normally seen by more than one window, at different offsets within each,
which is precisely why the model must tolerate off-centre echoes and why translation jitter during training
matters.

What happens to the model's output next is where the pipeline currently loses the most. Detections scoring below
the confidence threshold are discarded, which is expected. But then **only the highest-scoring detection in each
window is retained**, so two trucks falling in the same 64-pixel window cannot both be reported. The survivors
are then deduplicated greedily in descending score order, suppressing any detection whose red keypoint lies
within 32 pixels of one already kept — a radius of 96 metres on the ground, which merges genuinely adjacent
trucks on a freeway. Finally the red keypoint of each surviving detection is mapped back to map coordinates
through the raster's affine transform.

Both of those reductions were measured, and both are costly. With top-one-per-window and deduplication disabled
at a threshold of zero, every one of the eighty labelled trucks on a test scene has a detection within eighteen
metres. The model finds the trucks; the post-processing discards them.

# Outputs

For each window the model returns five parallel arrays, holding up to one hundred detections.

```{r outputs}
tibble(
  field = c("boxes", "labels", "scores", "keypoints", "keypoints_scores"),
  shape = c("(N, 4)", "(N,)", "(N,)", "(N, 3, 3)", "(N, 3)"),
  contents = c("box corners in resized-chip pixels",
               "class index, always 1",
               "box classifier confidence, 0 to 1",
               "x, y and visibility per keypoint, in blue-red-green order",
               "per-keypoint peak logit, unbounded"),
  `read by the pipeline` = c("no", "no", "yes", "yes", "no")
)
```

Three of the five are computed and thrown away. The boxes are never read, because matching is done on keypoint
distance rather than box overlap. The labels are uninformative given a single foreground class. The keypoint
confidences are discarded despite being the most discriminative signal the model produces, which is the largest
identified inefficiency in the current pipeline.

A hard floor also exists inside the model: detections scoring below 0.05 are dropped internally before any
application code sees them, so 0.05 is the lowest threshold recoverable without re-running inference.

One property of the confidence score has practical consequences worth stating explicitly. The threshold is a
filter applied after detection, not a parameter of detection — it changes which detections survive, not which are
found. Recording every detection together with its score therefore allows any higher threshold to be evaluated
afterwards by simple filtering, so a sweep of dozens of thresholds costs seconds rather than repeated inference
runs. Deduplication commutes with that filtering, because it proceeds in descending score order and a
lower-scoring detection can never suppress a higher-scoring one, so the reconstruction is exact rather than
approximate. This was verified across every scene and threshold combination in the dataset.

Beyond the raw model output, the collection pipeline persists derived quantities per detection: the three
keypoints in both pixel and map coordinates, the box and its dimensions, all four confidence values, and the echo
geometry — streak length in pixels and metres, the two segment lengths, their ratio, the perpendicular deviation
of the red keypoint from the line joining blue and green, and the bearing. Evaluation fields record the distance
to the nearest label, which label was matched, and the keypoint localisation error where a match occurred.

# Evaluation

A detection is judged correct if its red keypoint lies within six pixels — eighteen metres — of a labelled
vehicle's red keypoint, and each label may be claimed by at most one detection.

This departs from standard object-detection practice, which matches by bounding-box intersection-over-union.
Two reasons motivate the departure. A two-to-six-pixel echo has a box so small that IoU becomes dominated by
quantisation noise, and the red keypoint is the physically meaningful anchor, marking the vehicle's position at
the middle exposure. The consequence is that these figures are not directly comparable to published results:
Adamiak and colleagues report a mean average precision of 0.59 using COCO's object keypoint similarity, and Van
Etten reports an F1 of 0.49 using segmentation-based matching, and neither uses the metric applied here.

From the matching, true positives are detections within the radius, false positives are detections without a
nearby label, and false negatives are labels without a nearby detection. Precision, recall and their harmonic
mean follow conventionally. A count ratio — detections divided by labelled vehicles — is also tracked, since the
programme's actual deliverable is aggregate truck counts rather than per-vehicle identification. There is no true
negative: a window producing no detection generates no record, so accuracy and specificity are undefined for this
task.

Per-scene results are combined in two ways, and the distinction matters. Micro-averaging sums true positives,
false positives and false negatives across scenes before computing the metric once, which weights scenes by
vehicle count and lets large scenes dominate. Macro-averaging computes the metric per scene and then averages,
weighting every scene equally so that a two-vehicle scene counts as much as a hundred-vehicle one. Both are
reported, and a divergence between them indicates that performance correlates with scene size.

# Known limitations

The most consequential limitation is the labelling policy. Because only clear echoes are marked, the imagery
contains real vehicles that are deliberately unlabelled, so precision is a lower bound and every F1 figure
understates the model by an unknown margin. Establishing the true rate requires exhaustively labelling at least
one scene, which has not been done.

The training data has no background-only chips. Torchvision samples hundreds of negative regions inside every
chip, so the model does see abundant examples of "not a truck" — but because every chip is centred on a vehicle,
those negatives are all drawn from road and near-road terrain. Field interior, rooftops and bare soil never
appear as labelled background, which is a plausible cause of spurious detections over unfamiliar ground.

All imagery falls within a single spring window spanning April to June, so seasonal robustness is entirely
untested, and agricultural corridors in particular will look substantially different at other times of year.

Validation shares scenes with training, so best-epoch selection is biased. No scene in the current corpus is
genuinely out-of-sample for both trained models, which means no clean measurement of transfer to unseen ground
exists anywhere in the project. And the post-processing discards detections the model successfully made, so
reported recall understates the network's actual detection capability by a wide margin.

---

_Architecture and hyperparameters verified against the installed torchvision implementation and the project
source: `src/export_coco.py`, `src/train_detector.py`, `src/detect_scene.py`, `src/model_registry.py`,
`src/collect_inference.py`. Reference method: Adamiak et al. 2025, Int. J. Applied Earth Observation and
Geoinformation 142:104707._
