---
title: "Detection results"
subtitle: "Two trained models tested on 21 labelled scenes across the full confidence range"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    number_sections: true
    theme: flatly
    df_print: kable
    fig_width: 10
    fig_height: 6
    code_folding: hide
  word_document:
    toc: true
    toc_depth: 2
    fig_width: 9
    fig_height: 5.5
    df_print: kable
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

```{css, echo=FALSE, eval=knitr::is_html_output()}
.main-container { max-width: 1220px; }

table {
  width: 100%;
  margin: 1.4em 0 1.8em 0;
  font-size: 0.92em;
  border-collapse: collapse;
}
table th, table td {
  padding: 7px 16px !important;
  white-space: nowrap;
  text-align: right;
}
table th:first-child, table td:first-child,
table th:nth-child(2), table td:nth-child(2) { text-align: left; }
table th {
  border-bottom: 2px solid #d5dbe0;
  font-weight: 600;
  color: #2c3e50;
}
table tbody tr:nth-child(even) { background: #f7f9fa; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 150, fig.retina = 2)
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(tidyr); library(tibble)
  library(ggplot2); library(forcats); library(scales)
})
MODELS <- c(MODEL_A = "adamiak-v2", MODEL_B = "warmup-v1")
PAL    <- c("adamiak-v2" = "#2a78d6", "warmup-v1" = "#eb6834")
FOCUS  <- c(0.3, 0.5, 0.7, 0.9)

theme_set(theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(colour = "grey35", size = 10),
        plot.caption = element_text(colour = "grey45", size = 9, hjust = 0),
        strip.text = element_text(face = "bold"),
        legend.position = "top", legend.title = element_blank()))

results <- read_csv("data/results.csv", show_col_types = FALSE) |>
  mutate(model = recode(model, !!!MODELS), model = factor(model, levels = unname(MODELS)))
sweep <- read_csv("data/threshold_sweep.csv", show_col_types = FALSE) |>
  mutate(model = recode(model, !!!MODELS), model = factor(model, levels = unname(MODELS)))
scenes <- read_csv("data/scenes.csv", show_col_types = FALSE)

pooled <- function(df, ...) {
  df |> group_by(...) |>
    summarise(TP = sum(TP), FP = sum(FP), FN = sum(FN),
              detected = sum(detected_count), annotated = sum(annotated_count), .groups = "drop") |>
    mutate(precision = TP / pmax(TP + FP, 1), recall = TP / pmax(TP + FN, 1),
           f1 = 2 * TP / pmax(2 * TP + FP + FN, 1), count_ratio = detected / pmax(annotated, 1))
}
foc <- results |> filter(threshold %in% FOCUS)
```

# Where the data comes from

## The imagery

Every scene is PlanetScope imagery from Planet Labs, licensed for education and research. The product is a
`PSScene` in the `analytic_8b_sr_udm2` bundle from the PSB.SD "SuperDove" instrument: eight colour bands,
unsigned 16-bit, corrected to Surface Reflectance, at three metres per pixel.

Each image was ordered as a clip around a freight corridor rather than as a full satellite footprint, which keeps
file sizes down and puts the coverage on road. Scenes were ordered with reprojection turned off, so each arrives
on its native grid in its native UTM zone: zone 10 for western Washington, zone 11 for the east. That matters,
because the detector relies on a very small offset between colour bands, and resampling the image would blur that
offset away. The imagery is never re-gridded at any point.

There are **24 scenes covering 205 square kilometres**, taken between **9 April and 22 June 2026**. Twenty-one of
them are labelled and make up the test set. The other three have imagery but no labels, and are left out of every
number in this report.

The education licence caps how much area can be ordered each month and bills by whole scene rather than by the
area actually used, so orders were kept small and targeted. That is why some corridors have much more coverage
than others, and why all the imagery comes from one spring period rather than spanning the year.

## The corridors

Scenes were placed on Washington freight corridors, favouring locations near WSDOT permanent traffic counters so
that model output could later be checked against measured traffic volumes.

```{r corridor-source}
scenes |> filter(labelled_vehicles > 0) |>
  group_by(Corridor = corridor) |>
  summarise(Scenes = n(), Vehicles = sum(labelled_vehicles),
            `Area (km²)` = round(sum(valid_km2), 1),
            `Vehicles per km²` = round(sum(labelled_vehicles) / sum(valid_km2), 2),
            `Date range` = if (n() == 1) as.character(min(date))
                           else paste(min(date), "to", max(date)), .groups = "drop") |>
  arrange(desc(Vehicles))
```

Coverage is very uneven, and that affects everything else in this report. Centralia provides 281 of the 789
labelled vehicles, over a third of the total, from only 13 square kilometres, because it is a busy stretch of
Interstate 5 carrying heavy freight. The dry eastern corridors are the opposite: Yakima covers 51 square
kilometres, four times Centralia's area, and has fewer than half as many vehicles. Traffic density ranges from
about 21 vehicles per square kilometre on Centralia to 1.2 around Ellensburg, a difference of roughly eighteen
times.

That imbalance reflects the real road network rather than a mistake in sampling, but it means any average
weighted by vehicle count is driven mostly by one corridor. This is why the report gives both vehicle-weighted
and scene-weighted averages.

## The labels

All 789 vehicles were labelled by hand in QGIS, working on the GeoTIFF in true colour so that labels carry
real-world coordinates rather than pixel positions. Each vehicle is marked with three points, one on the blue
blob, one on the red and one on the green, each tagged with the number of the exposure it belongs to.

Labelling was deliberately cautious. Only clear, well-formed echoes were marked, and blurry, misshapen,
over-bright and very small ones were skipped. **This affects every number in this report.** The imagery contains
real vehicles that were knowingly left unlabelled, so a detection that lands on one of them is counted as wrong
even though it is right. Precision as reported is therefore a floor, not an estimate, and both models are better
than the numbers below. How much better could only be settled by labelling one scene completely.

## The two models

Both are Keypoint R-CNN detectors with the same architecture. They differ in what they were trained on and how
training was scheduled.

`adamiak-v2` was trained on nine scenes containing 228 chips, 78 percent of them from Centralia. `warmup-v1` was
trained later on twenty scenes and 624 chips, with Centralia reduced to 23 percent, and added learning-rate
warmup, translation-jitter augmentation and multi-vehicle chip targets.

Because the training sets differ so much, every comparison in this report also reflects the difference in
corridor coverage. That means neither model can be shown to be better because of the training changes alone. What
can be shown is that the two behave differently, in ways that are consistent and repeatable.

## How the test was run

Both models were run over all 24 scenes with the same settings: a 64-pixel window moving in steps of 40 pixels,
detections kept only above the confidence threshold, duplicates within 96 metres removed, and a detection counted
as correct when its red keypoint lands within 18 metres of a labelled vehicle.

Each scene was processed once at a threshold of zero, and every detection was saved with its score. The threshold
filters detections after the model has run rather than changing how it runs, so any higher threshold can be
worked out by filtering those saved records. That is how the report covers 28 thresholds without running the
detector more than once per scene.

# What is being measured

The detector does not recognise trucks by shape. At three metres per pixel a truck is only a few pixels across.
It works instead from the way the satellite records its colour bands a fraction of a second apart, which makes a
moving vehicle appear as a short blue, red and green streak, called a moving echo. A parked vehicle does not
streak, so only moving traffic is detected.

The model reports each detection with a **confidence score** between 0 and 1. A **confidence threshold** decides
which detections are kept: raise it and fewer, more certain detections survive. The threshold is applied after
the model has run, so it can be changed without running the detector again. It is the main setting in this
report, and it changes the results more than the difference between the two models does.

Each kept detection is compared against the hand-labelled vehicles. If it lands within 18 metres of one, it
counts as correct. From those matches come the four numbers used throughout:

**Precision** is the share of detections that are real. **Recall** is the share of real vehicles that were found.
**F1** combines the two into one number that is only high when both are high. **Count ratio** is simply the
number of detections divided by the number of labelled vehicles, so 1.0 means the right total was reported,
whether or not the individual vehicles were the right ones. Count ratio is reported separately because the goal
of the project is traffic counts, and a model can report the right total while getting many individual vehicles
wrong.

Two of these appear in two forms. **Pooled** figures add the counts across all 21 scenes and work out the metric
once, which gives every vehicle equal weight and lets the busy corridors dominate. **Per-scene** figures work out
the metric for each scene and then average, which gives every scene equal weight. Both are reported because
coverage is so uneven, and the size of the gap between them tells you how much a model's headline number
depends on a few busy scenes.

Full definitions, including how each number is calculated and what the model actually produces, are in the
glossary at the end of the report.

# Which confidence levels this report compares

The confidence threshold sets how choosy the detector is. Low values keep nearly everything it proposes; high
values keep only what it is most sure about. It is chosen after training rather than learned, and it changes the
results more than the difference between the two models does. This report compares four evenly spaced levels,
**0.3, 0.5, 0.7 and 0.9**, covering the range from loose to strict.

**0.3** is the loose end, close to what the detector proposes with almost no filtering. It shows how much each
model can find at best, and how many wrong detections come with that.

**0.5** is the current setting, the value the inference console defaults to and the setting behind every number
previously reported in this project.

**0.7** is the strict but still usable middle. Model B is near its best here, and it is where the two models come
closest to each other.

**0.9** is the strict end, where only near-certain detections survive. A counting programme might reasonably
prefer to trust what it reports over finding everything, and this level shows what that costs.

The best settings do not fall on these round numbers: model A peaks at 0.55 and model B at 0.65. Both are marked
on the full sweep later in the report, which covers all 28 thresholds rather than these four.

# Results at the four confidence levels

## Overall performance, pooled across all scenes

The table below is **pooled**. Every scene's true positives, false positives and false negatives are added
together across the 21 scenes, and precision, recall and F1 are worked out once from those totals. This gives
every vehicle equal weight, so the busy corridors dominate; Centralia alone accounts for 281 of the 789 labelled
vehicles. It answers the question the project actually cares about: across the whole area surveyed, what share of
trucks were found and what share of detections were real.

```{r headline}
pooled(foc, model, threshold) |>
  transmute(Model = model, Threshold = threshold, TP, FP, FN,
            Detections = detected, Labelled = annotated,
            Precision = round(precision, 3), Recall = round(recall, 3),
            F1 = round(f1, 3), `Count ratio` = round(count_ratio, 2))
```

At **0.3** both models report far too much. Model A returns 1,218 detections against 789 labelled vehicles, and
model B returns 2,252, nearly three times the true count. Recall is at its highest here, 53 percent for model A
and 60 percent for model B, but precision is poor at 34 and 21 percent. Neither is usable as a count at this
setting. The recall figures are still useful, because they are the closest thing in the table to what each model
can find at all.

At the **current setting of 0.5** the two separate. Model A tightens to 708 detections, just under the 789
present, and reaches its best F1 in this table at 0.480, with 51 percent precision against 46 percent recall.
Model B still returns 1,603 detections, twice the true count. Its F1 of 0.355 is caused by the volume of wrong
detections rather than by failing to find trucks: its recall, at 54 percent, is higher than model A's.

At **0.7** the two almost meet, and this is the most useful row in the table. Model B improves to 0.376, its best
of the four. Model A drops to 0.441, because it is past its best setting near 0.55 and the filter is now removing
more real detections than wrong ones. Their per-scene averages are effectively tied, 0.351 for model A against
0.352 for model B. Model B's count is also at its closest to correct here, 1,044 detections against 789 vehicles.

At **0.9** the order reverses. Model A becomes very selective, at 87 percent precision, the highest in this
table, but it finds only 14 percent of the vehicles and its F1 drops to 0.237. Model B holds up much better at
0.313 and is now clearly ahead. On some corridors model A reports nothing at all at this level: it scores zero on
Stanwood, having found none of that scene's vehicles with enough confidence to report them.

No single threshold suits both models. Model A is at its best near 0.5 and is not usable by 0.9. Model B is at
its best near 0.7 and still works at 0.9.

## Variation between scenes

The overall figures hide how differently the two models behave from one scene to the next, which for a monitoring
programme matters as much as the average.

```{r box-helper, include=FALSE}
box_metric <- function(var, title, sub, cap) {
  foc |>
    ggplot(aes(factor(threshold), .data[[var]], colour = model, fill = model)) +
    geom_boxplot(alpha = 0.16, width = 0.5, outlier.shape = NA, linewidth = 0.6,
                 position = position_dodge(width = 0.66)) +
    geom_point(position = position_jitterdodge(jitter.width = 0.13, dodge.width = 0.66),
               alpha = 0.65, size = 2) +
    scale_colour_manual(values = PAL) + scale_fill_manual(values = PAL) +
    scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
    labs(title = title, subtitle = sub, x = "confidence threshold", y = NULL, caption = cap)
}
```

```{r fig-box-precision, fig.height=5, fig.width=9}
box_metric("precision", "Precision by scene",
  "Each point is one of the 21 labelled scenes; boxes show the middle half of the range.",
  "Precision rises with the threshold for both models. Model A varies far more between scenes, especially at 0.9.")
```

Precision is where the two models differ most. Model A is the more precise at every level, and the gap grows as
the threshold rises, from 0.34 against 0.22 at 0.3 to 0.70 against 0.42 at 0.9. But its results vary across
almost the whole range from scene to scene, and at 0.9 that variation is extreme, with a standard deviation of
0.39. This happens because model A returns so few detections on quiet corridors at 0.9 that one hit or miss moves
that scene's precision by tens of points, leaving scenes at 100 percent next to scenes at zero. Model B's
precision is lower throughout but far steadier, staying close to its median on nearly every scene.

```{r fig-box-recall, fig.height=5, fig.width=9}
box_metric("recall", "Recall by scene",
  "Each point is one of the 21 labelled scenes; boxes show the middle half of the range.",
  "Recall falls with the threshold for both models. Model B is higher and steadier at every level.")
```

Recall separates the two most clearly. Model B is above model A at every level and stays in a narrow band, with a
standard deviation near 0.10 throughout, while model A varies much more and falls away faster. The two are
closest at 0.3, where model A still finds half the vehicles, and furthest apart at 0.9, where model B finds a
quarter of them and model A one in ten. This is model B's wider training coverage showing through: it finds a
similar share of vehicles almost everywhere, while how well model A does depends heavily on which corridor it is
looking at.

```{r fig-box-f1, fig.height=5, fig.width=9}
box_metric("f1", "F1 by scene",
  "Each point is one of the 21 labelled scenes; boxes show the middle half of the range.",
  "Model A leads at 0.3 and 0.5, the two are tied at 0.7, and model B leads at 0.9. Model B varies less throughout.")
```

F1 puts the two together, and the four levels show a clear crossover. Model A leads at 0.3 and 0.5, the two are
effectively tied at 0.7 with 0.351 against 0.352, and model B is clearly ahead at 0.9 with 0.289 against 0.172.
Model A's lead at the lower levels comes with a run of poor scenes that pulls its average well below its median,
while model B's worst scenes are not far from its best. Which model looks better depends on where the threshold
is set.

The table below summarises this. It is **not** a repeat of the pooled table above. Here each scene's precision,
recall and F1 are worked out separately and then averaged across the 21 scenes, giving every scene equal weight
instead of every vehicle. A six-vehicle scene therefore counts as much as Centralia, which is why these averages
sit below the pooled ones: at 0.5 model A reads 0.402 mean F1 here against 0.480 pooled, because its weakest
results come from quiet scenes that the pooled figure barely registers. The standard deviation and variance
columns are the point of this table, and they have no pooled equivalent.

```{r spread-table}
foc |>
  group_by(Model = model, Threshold = threshold) |>
  summarise(`Scenes` = n(), `F1 mean` = round(mean(f1), 3), `F1 median` = round(median(f1), 3),
            `F1 SD` = round(sd(f1), 3), `F1 var.` = round(var(f1), 4),
            `Recall mean` = round(mean(recall), 3), `Recall SD` = round(sd(recall), 3),
            `Prec. mean` = round(mean(precision), 3), `Prec. SD` = round(sd(precision), 3),
            .groups = "drop")
```

Model B is clearly the steadier of the two. Its per-scene F1 standard deviation runs from 0.124 down to 0.094
across the four levels, against model A's 0.195 to 0.125, and its variance is close to forty percent of model A's
at 0.3, 0.5 and 0.7. The pattern is stronger in recall, where model B stays in a narrow band on nearly every
scene while model A ranges from eighteen percent recall on Spokane to sixty-seven percent on Stanwood and
auburn-snoqualmie at the default threshold. Model A only becomes steadier at 0.9, and for the wrong reason: a
model that reports almost nothing is consistently bad rather than reliably good.

Part of the reason shows in the scene-level results. At 0.5 model A reaches a pooled F1 of 0.526 on the nine
scenes it trained on against 0.454 on the twelve it did not. That is a real gap, and one model B cannot be tested
for, since only a single scene was kept back from it. But the pattern is not clean. Model A's best corridor,
auburn-snoqualmie at 0.667, is one it never trained on, while two of its weakest, Stanwood and Ellensburg, are
corridors it did train on. Training coverage explains some of model A's variation but not all of it. What can be
said plainly is that model A is not the better detector everywhere; it is the less predictable one.

This matters in practice. The goal is total truck counts, and a detector that is wrong by a similar amount
everywhere can be corrected with one adjustment factor. A detector that is good in some places and nearly blind
in others needs a different correction in each place, which is much harder to justify.

## Counting behaviour

Because the goal is counts rather than identifying individual vehicles, how many detections each setting produces
compared to the true number matters as much as whether each one is right.

```{r fig-count, fig.height=9, fig.width=11}
YCAP  <- 8
above <- foc |> filter(count_ratio > YCAP)

foc |>
  ggplot(aes(factor(threshold), count_ratio, colour = model, fill = model)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey25", linewidth = 0.7) +
  geom_boxplot(alpha = 0.16, width = 0.56, outlier.shape = NA, linewidth = 0.6,
               position = position_dodge(width = 0.68)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.14, dodge.width = 0.68),
             alpha = 0.65, size = 2) +
  annotate("text", x = 4.48, y = 1, label = "count parity", vjust = -0.7, hjust = 1,
           size = 3.4, colour = "grey25") +
  scale_colour_manual(values = PAL) + scale_fill_manual(values = PAL) +
  scale_y_continuous(breaks = seq(0, YCAP, 1), expand = expansion(mult = c(0.02, 0.06))) +
  coord_cartesian(ylim = c(0, YCAP)) +
  labs(title = "Detections compared to labelled vehicles",
       subtitle = paste0("Linear scale. The dashed line is count parity: one detection per labelled vehicle. ",
                         nrow(above), " points fall above the axis (highest ",
                         sprintf("%.1f", max(above$count_ratio)),
                         "), all from EllensburgPreferredTest_01, a scene with only 2 labelled vehicles."),
       x = "confidence threshold", y = "detections ÷ labelled vehicles",
       caption = "Model A reaches parity near 0.45, model B near 0.775. Boxes use all the data; the axis is cut off at 8 for readability.")
```

The four levels sit either side of count parity for both models, but at very different points. The ratios given
here are pooled across the network. The per-scene points in the figure vary much more, because a scene with only
a few labelled vehicles gives a very large ratio from a small error. The four points above the top of the axis
all come from one two-vehicle scene, where 20 to 49 detections against two labels gives ratios between 10 and
24.5. That says something about the small number of labels, not about the scene being hard.

Model A reports about half again too many at 0.3, sits just under parity at 0.5 with 0.90 detections per labelled
vehicle, and then reports too few: about half the true number at 0.7 and a sixth at 0.9. Model B reports too many
at every level up to 0.7, nearly three times too many at 0.3 and twice too many at 0.5, comes closest at 0.7 with
1.32, and passes through parity between 0.7 and 0.9. Neither counts correctly at the current setting: model A is
10 percent low and model B is twice high.

The threshold that gives an accurate count is not the one that gives the best F1. Model A reaches parity near
0.45 but peaks on F1 near 0.55. Model B reaches parity near 0.775 but peaks near 0.65.

Some of the over-counting comes from the cautious labelling rather than from real error, since unlabelled real
vehicles raise the detection count without raising the number they are compared against. This affects the
higher-recall model more, so model B's true ratios are lower than the plotted ones by an unknown amount.

## How performance varies by corridor

```{r fig-corridor, fig.height=6.5, fig.width=10}
pooled(foc, corridor, model, threshold) |>
  mutate(corridor = fct_reorder(corridor, f1, .fun = max)) |>
  ggplot(aes(f1, corridor, fill = model)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.7) +
  facet_wrap(~ paste("threshold", threshold), nrow = 1) +
  scale_fill_manual(values = PAL) +
  labs(title = "F1 by corridor at each threshold",
       subtitle = "Pooled within corridor, at each of the four confidence levels.",
       x = "F1", y = NULL,
       caption = "Model A is strongest on auburn-snoqualmie and Centralia; model B is more even across the network.")
```

Performance follows traffic density more than terrain, and the two models fail in different places. At the
default threshold model A is strongest on auburn-snoqualmie, Tri-Cities and Centralia, at 0.657, 0.622 and 0.597,
which are the busy interurban corridors. It does much worse on the quiet ones, reaching only 0.256 on Spokane,
0.264 on Ellensburg and 0.267 on Kent/Auburn, where a few labelled vehicles spread over a large area make every
error costly. Model B covers a much narrower range at the same threshold, from 0.428 on Centralia down to 0.200
on Kent/Auburn.

The dry eastern corridors separate the two most, and in both directions. Model B beats model A on Spokane, 0.423
to 0.256, and on Ellensburg, 0.325 to 0.264. Model A wins Tri-Cities clearly, 0.622 to 0.401, and is slightly
ahead on Yakima. Model A's poor results on the quiet corridors are the main reason its scene-weighted average
falls below its vehicle-weighted one.

At 0.9 the picture changes completely. Model A stays usable only on the three busy corridors and drops below 0.15
on five of the nine, including zero on Stanwood. Model B stays above 0.24 on every corridor, and its best at this
level is Spokane at 0.444, a corridor model A never trained on and scores 0.102 on. Model B's best corridor
result across all four levels is close to that: 0.449, at 0.7 on both Spokane and Ellensburg.

# Behaviour across the full confidence range

The four levels above are samples from a continuous range. Sweeping the whole range shows how each model responds
and finds the setting where each does best, including the best settings that fall between the round numbers.

```{r fig-sweep, fig.height=5.4, fig.width=11}
peaks <- sweep |> group_by(model) |> slice_max(mean_F1, n = 1) |> ungroup()
sweep |>
  ggplot(aes(threshold, mean_F1, colour = model)) +
  geom_vline(xintercept = FOCUS, colour = "grey88") +
  geom_line(linewidth = 1) +
  geom_point(data = peaks, size = 3.4, shape = 21, stroke = 1.5, fill = "white") +
  geom_text(data = peaks, aes(label = sprintf("peak %.3f at %.2f", mean_F1, threshold)),
            vjust = -1.4, size = 3.3, show.legend = FALSE) +
  scale_y_continuous(limits = c(0, 0.45)) +
  scale_colour_manual(values = PAL) +
  labs(title = "Mean F1 across the confidence range",
       subtitle = "Averaged over the 21 labelled scenes. Grey lines mark the four levels compared above.",
       x = "confidence threshold", y = "mean F1",
       caption = "Each curve has one peak: F1 rises while the filter is removing wrong detections, then falls once it starts removing real ones.")
```

Both curves rise, peak and fall. At first the filter removes more wrong detections than real ones, so results
improve. Past the peak that reverses and real detections are lost faster than wrong ones. **Model A peaks at 0.55
with a mean F1 of 0.403, and model B at 0.65 with 0.361.** Neither was being run at its best setting, and the
current default of 0.5 sits below both.

```{r fig-sweep-prf, fig.height=4.8, fig.width=11}
sweep |>
  select(model, threshold, Precision = pooled_P, Recall = pooled_R, `Count ratio` = count_ratio) |>
  pivot_longer(c(Precision, Recall, `Count ratio`), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric, levels = c("Precision", "Recall", "Count ratio"))) |>
  ggplot(aes(threshold, value, colour = model)) +
  geom_hline(data = tibble(metric = factor("Count ratio",
              levels = c("Precision", "Recall", "Count ratio")), y = 1),
             aes(yintercept = y), linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 1) +
  facet_wrap(~ metric, nrow = 1, scales = "free_y") +
  scale_colour_manual(values = PAL) +
  labs(title = "Precision, recall and count ratio across the confidence range",
       subtitle = "Precision rises steadily, recall falls steadily, and the count ratio drops through parity.",
       x = "confidence threshold", y = NULL,
       caption = "Each model reaches count parity at a different threshold: model A near 0.45, model B near 0.775.")
```

Precision and recall move in opposite directions across the whole range, which is why F1 peaks somewhere in the
middle. The count ratio falls steadily and passes through parity at a different threshold from the F1 peak, near
0.45 for model A and 0.775 for model B. The setting that gives the best detection quality is therefore not the
setting that gives the most accurate count.

This is also why very high thresholds were ruled out. At 0.95 model B finds 17 percent of the labelled vehicles
and model A about 5 percent, and both score worse on F1 than they do at 0.3, model A far worse at 0.094 against
0.417. Raising the threshold does stop the over-counting, but it replaces it with heavy under-counting.

# Summary

**Model A reaches the higher F1.** Its best measured setting is a threshold near 0.55, giving a mean per-scene F1
of 0.403 and a pooled F1 of 0.485. For comparison, Van Etten's 2024 study of vehicle detection in PlanetScope
imagery reports an F1 of 0.49 on a similar task, so the detector is working at roughly the level published work
reaches at this resolution. The two studies define their metrics differently enough that this is a rough
comparison rather than an exact one.

**Model B is the more consistent detector.** Its per-scene F1 varies about forty percent less than model A's, and
its recall stays within a narrow band across every corridor, including the ones it never trained on. Model A is
better on the busy corridors and much worse on the quiet ones, which is what produces its higher average and its
much wider spread.

Which of those two properties is worth more depends on the use. For a single accurate count of a busy stretch of
road, model A is the better choice. For anything that has to work the same way across the whole network, model B
is, because a detector that is wrong by a similar amount everywhere can be corrected with one adjustment factor,
while one that is good in some places and blind in others cannot.

**Neither model is in good enough shape to estimate annual truck counts from new imagery.** Both need a threshold
chosen for them, and the setting that gives the best F1 is not the setting that gives the right count. Neither
counts correctly at the current default: model A reports 10 percent too few, model B twice too many. More
importantly, nothing here shows how either model behaves on ground it has not seen, or at times of year it has
not seen, and both of those are required before a detection count can be turned into an annual traffic estimate.

# Limitations

**Precision is a floor, not an estimate.** Only clear echoes were labelled, and the imagery contains real
vehicles that were left unmarked, so both models are better than these numbers show. The gap cannot be measured
without labelling one scene completely.

**There is no clean test of how the models handle new ground.** Only one scene was kept back from both models,
and its footprint overlaps ground they trained on by 97 percent, so it is a repeat visit to familiar terrain
rather than a test of somewhere new.

**The two models cannot be compared cleanly.** Their training sets differ in both size and mix of corridors, so
the differences between them cannot be attributed to the training changes alone.

**All the imagery comes from one spring period**, between April and June, so nothing here shows how the detector
behaves in other seasons. That matters most for the farming corridors, which change appearance a great deal
through the year.

**The matching rule is specific to this project.** A detection counts when its red keypoint lands within 18
metres of a label, rather than the box-overlap test normally used in object detection, so comparisons to
published results are approximate.

Taken together, these points set the limit on what can be claimed. The results show that the method works and
that model A is more accurate while model B is more consistent. They do not show that either model can be pointed
at new imagery and produce a reliable annual count.

# What goes into the model and what comes out

This section lists everything the detector is given and everything it produces, in the order it moves through
the pipeline. It is a reference for anyone who needs to reproduce or extend the work.

## Inputs: the imagery

Each scene starts as a GeoTIFF with eight colour bands, unsigned 16-bit, at three metres per pixel. Only three
bands are used. Band 6 becomes the red channel, band 4 the green and band 2 the blue, which gives a true-colour
image and is also the order the sensor records them.

Those bands are 16-bit, and almost all the real values sit in a narrow part of that range, so each band is
stretched before use: the 2nd and 98th percentile of the non-zero pixels are mapped onto 0 to 255. The
percentiles are worked out separately for each scene, so two scenes are not stretched by the same amounts. Pixels
with no data stay black.

The stretched image is then covered by a sliding 64-pixel window moving in steps of 40 pixels. A window is only
processed if at least 15 percent of its pixels contain data, which skips the empty area outside the ordered clip.
Each window is passed to the model as a 3 × 64 × 64 array of values divided by 255. Inside the model, torchvision
resizes it to a minimum side of 192 pixels and normalises it using the standard ImageNet mean and standard
deviation.

## Inputs: the labels

Labels live in a GeoPackage layer with one row per point, so three rows per vehicle. Each row carries the point
geometry in the scene's coordinate system, a `vehicle_id` grouping the three points, a `sequence` number saying
which band the point marks (1 blue, 2 red, 3 green), and a `scene` text field.

That `scene` field is what joins labels to imagery. The join is never done by location, because several scenes
overlap on the ground and a location-based match would put one scene's labels on another scene's pixels. Only
vehicles that have all three sequence numbers are used; incomplete ones are dropped.

Labels are used for two separate purposes: to build training chips, and to score detections. No scene is used
for both purposes in the same model.

## Inputs: model configuration

Each trained model stores its own configuration, because the settings below change the shape of the network and
weights saved under one set will not load under another.

| Setting | Value |
|---|---|
| Backbone | ResNet-50 with Feature Pyramid Network |
| Classes | 2 (moving echo, background) |
| Keypoints | 3 (blue, red, green) |
| Anchor sizes | 4, 8, 16, 32, 48 pixels |
| Anchor aspect ratios | 0.25, 0.5, 0.75, 1.0, 1.25 |
| Input resize | minimum side 192, maximum side 320 |

## Inputs: run settings

These are chosen at run time and do not change the model. They are the settings that produced every number in
this report.

| Setting | Value | What it does |
|---|---|---|
| Window size | 64 px | Size of each view passed to the model |
| Stride | 40 px | Step between windows, so they overlap by 24 px |
| Minimum valid fraction | 0.15 | Skips windows that are mostly empty |
| Confidence threshold | 0.3 to 0.95 | Which detections are kept |
| Duplicate radius | 32 px (96 m) | Minimum spacing between kept detections |
| Match radius | 6 px (18 m) | How close a detection must be to a label to count |
| Ground sample distance | 3 m | Used to convert pixels to metres |

Three further settings are internal to torchvision and were left at their defaults: a region score cutoff of
0.05, a non-maximum-suppression overlap of 0.5, and a cap of 100 detections per window.

## Outputs: what the model returns for each window

For every window, the model returns five arrays, where N is the number of candidate detections it found in that
window, ordered from highest score to lowest.

| Output | Shape | What it is | Used here |
|---|---|---|---|
| `scores` | (N) | Confidence that the region holds a moving echo | Yes |
| `keypoints` | (N, 3, 3) | x, y and a visibility flag for each of the three keypoints | Yes, x and y |
| `boxes` | (N, 4) | The rectangle around each detection | Recorded, not used |
| `keypoints_scores` | (N, 3) | How sharp each keypoint's heatmap peak was | Recorded, not used |
| `labels` | (N) | Class number, always 1 here | No |

The pipeline keeps only the highest-scoring detection in each window, then drops any detection whose red
keypoint falls within 96 metres of a higher-scoring one already kept. What remains is a detection, with a
confidence score, three keypoint positions in scene pixels, and the red keypoint converted to UTM easting and
northing using the scene's geotransform.

## Outputs: files written per scene

Running the detector on one scene writes a JSON file of detections, each with its score, its three keypoints in
pixels and its red keypoint in UTM. It also writes a preview image of the whole scene and a montage of close-up
crops around each detection for visual checking. If the scene is labelled, it also reports how many labelled
vehicles were matched, how many detections landed near a label and how many did not.

## Outputs: the analysis tables

The results in this report come from four tables built from those per-scene runs.

**Per-detection records**, one row per surviving detection, are the most detailed output. Each row holds which
model and scene it came from and whether that scene was trained on or labelled; the window it was found in and
whether it was the top-scoring detection there; its confidence; the box position and size; all three keypoints in
pixels and in UTM; the three keypoint scores and their mean; measured streak geometry, meaning the streak length
in pixels and metres, the two segment lengths, their ratio, how straight the three points are and the compass
bearing; and finally its distance to the nearest label, which label it matched if any, and the keypoint error in
pixels. There are 44 columns in total, covering roughly 21,000 detections per model.

**Per-scene results** hold one row per scene, model and threshold: the scene and corridor, the threshold, the
number of labelled vehicles and detections, true positives, false positives and false negatives, and the derived
precision, recall, F1, count ratio and signed count error. This table has 336 rows and is what every figure and
table in this report is calculated from.

**The threshold sweep** holds one row per model and threshold across 28 thresholds, with mean and standard
deviation of per-scene F1, and the pooled F1, precision, recall and count ratio.

**Scene metadata** holds one row per scene with its corridor, date, coordinate system, size in pixels, valid area
in square kilometres, number of windows processed and number of labelled vehicles.

Everything in this report is derived from those four tables. No number was entered by hand.

# Glossary and definitions

Definitions for the terms and metrics used above, including how each is calculated. Ordinary statistics such as
means and standard deviations are used in the normal way and are not defined here.

## The moving echo

The signal the detector looks for. The SuperDove sensor records its colour bands a fraction of a second apart. A
parked object sits in the same place in every band, but a moving vehicle has shifted between them, so combining
three bands into a colour image shows it as a short streak of separated colour blobs instead of a single object.
In the band combination used here the streak runs blue, red, green, which is the order the sensor records those
bands rather than their numbering. Everything in this report finds and measures that streak, not the vehicle
itself.

## Keypoints

A **keypoint** is a predicted point marking where one band's image of the vehicle sits. Every vehicle, whether
labelled by hand or found by the model, is represented by three of them: blue, red and green. Together they give
the streak's position, length and direction.

The model does not predict these coordinates directly. For each region it looks at, the keypoint head produces
three heatmaps $H_1, H_2, H_3 \in \mathbb{R}^{56 \times 56}$, one per keypoint, over a grid laid across that
region. The predicted position of keypoint $k$ is the cell with the highest value,

$$(u_k, v_k) = \arg\max_{(u,v)} H_k(u,v),$$

which is then converted back to scene pixel coordinates using the region's position and scale, and from there to
UTM easting and northing using the scene's geotransform. Training handles this as a choice between cells rather
than as a measurement: the $56 \times 56$ grid is flattened into 3,136 possible positions and the model is scored
on whether it picked the right one. It is asked which cell holds the band, not how far off its guess was.

Three uses matter later. The **red keypoint** is treated as the vehicle's position, and every distance in this
report is measured between red keypoints. The distance from blue to green gives the **streak length**, which
grows with speed. The spacing and straightness of the three points describe the shape of the streak, which is
what separates a real echo from a bright patch of ground.

## Keypoint score

The model also reports the heatmap value at each chosen peak, $H_k(u_k, v_k)$. This is a raw score rather than a
probability: it has no fixed range, it is not calibrated, and in this data it runs from about $-4$ to $+8$. It
measures how sharp the peak is, so a clear band position scores high and a vague one scores low or negative. The
pipeline that produced these results saves these values but does not filter on them, so they change nothing in
this report. They are defined here because they appear in the detection data files.

## Confidence, and the confidence threshold

**Confidence**, also called the detection score, is the model's estimate that a region holds a moving echo rather
than background. For each region the classification head produces two values, $z_0$ for background and $z_1$ for
the echo class, and the confidence is their softmax:

$$s = \frac{e^{z_1}}{e^{z_0} + e^{z_1}} \in (0, 1).$$

So it falls between 0 and 1 and behaves like a probability, but it is not calibrated. A detection scoring 0.9 is
not right 90 percent of the time. Only the ranking can be relied on: higher-scoring detections are more likely to
be real than lower-scoring ones. That is all a threshold needs.

The **confidence threshold** $t$ is the cutoff applied to that score. Detections with $s \ge t$ are kept and the
rest are dropped. It is applied after the model has run and is not something the model is trained with. The model
behaves the same way at every threshold; only the number of detections that survive changes. This is why one pass
over a scene, saved at threshold 0, covers every threshold in this report, and why the sweep can cover 28 values
without running the detector again.

## Windows, stride and duplicate removal

The detector was trained on 64-pixel chips, so a full scene is covered by sliding a **64-pixel window** across it
in steps of **40 pixels**. That step, the **stride**, is shorter than the window, so neighbouring windows overlap
by 24 pixels. The overlap makes sure no vehicle sits on a window edge, but it also means the same vehicle is
shown to the model several times and can be found more than once.

Two rules handle that. Only the highest-scoring detection in each window is kept, and then **duplicate removal**
drops any remaining detection whose red keypoint falls within **32 pixels (96 metres)** of a higher-scoring one
already kept. What survives both rules is what this report calls a **detection**.

## Matching a detection to a labelled vehicle

A detection counts as correct when its red keypoint falls within **18 metres**, which is 6 pixels at 3 metres per
pixel, of a labelled vehicle's red keypoint:

$$\lVert \mathbf{p}_{\text{red}}^{\text{det}} - \mathbf{p}_{\text{red}}^{\text{label}} \rVert_2 \le 18\ \text{m}.$$

This is a distance test rather than the box-overlap (IoU) test normally used in object detection. At 3 metres per
pixel a truck is only a few pixels across, so box overlap swings wildly on rounding alone, while the streak
position stays meaningful. Because of this, none of the numbers here can be compared directly to published mAP or
IoU results.

Duplicate removal keeps surviving detections at least 96 metres apart, and the match radius is 18 metres, so no
two detections can be matched to the same label. Each label is matched at most once.

## True positives, false positives and false negatives

A **true positive** (TP) is a labelled vehicle that has a matching detection. A **false positive** (FP) is a
detection that matches no label. A **false negative** (FN) is a labelled vehicle with no matching detection.
There is no true negative in this task: the imagery is continuous ground rather than a fixed list of candidates,
so "correctly reporting nothing" is not something that can be counted. Any metric that needs true negatives,
including accuracy, cannot be used here.

One point applies throughout. Only clear, well-formed echoes were labelled, so the imagery contains real vehicles
that were left unmarked. A detection that lands on one of those is counted FP even though it is right, which
makes every precision and F1 number in this report a **floor** rather than an estimate.

## Precision, recall and F1

**Precision** is the share of detections that are real: when the model reports a truck, how often it is right.
**Recall** is the share of real vehicles that were found: of the trucks present, how many the model found. **F1**
combines the two into one number (their harmonic mean) that is only high when both are high. A model can reach
perfect recall by reporting everything, or near-perfect precision by reporting almost nothing, and F1 rules out
both.

$$P = \frac{TP}{TP + FP}, \qquad R = \frac{TP}{TP + FN}, \qquad F_1 = \frac{2PR}{P + R} = \frac{2\,TP}{2\,TP + FP + FN}.$$

Raising the threshold always pushes precision up and recall down. This is why F1 peaks somewhere in the middle of
the range, and why the choice of threshold is a real decision.

## Count ratio

Detections divided by labelled vehicles over the same area,

$$\text{count ratio} = \frac{D}{A},$$

where $D$ is the number of detections and $A$ the number of labelled vehicles. A value of 1 is **count parity**,
meaning the right number of vehicles was reported, whether or not they were individually the right ones. It is
reported separately from F1 because the goal of the project is traffic counts, and the two can disagree. A model
can hit parity while getting many individual vehicles wrong, with its misses and its false detections cancelling
out. Above 1 means too many detections, below 1 too few.

## Pooled versus per-scene averaging

Every metric can be worked out two ways, and this report uses both because they answer different questions.

**Pooled** (micro-averaged) numbers add up the counts across all 21 scenes and work out the metric once from
those totals, for example $P = \sum_s TP_s / \sum_s (TP_s + FP_s)$. This gives every vehicle equal weight, so
busy corridors dominate. It answers what share of all the trucks across the network were found.

**Per-scene** (macro-averaged) numbers work out the metric for each scene separately and then average those
results, $\bar{P} = \frac{1}{S}\sum_s P_s$. This gives every scene equal weight, so a scene with six vehicles
counts as much as one with eighty. It answers how the detector does on a typical scene.

The two differ here because the models do better where traffic is heavier: model A reads 0.480 pooled F1 against
0.402 per-scene at the default threshold. The bigger that gap is, the more a model's headline number depends on a few busy scenes.

## Scenes, corridors, and training status

A **scene** is one satellite image of one place on one date, and is the unit of both imagery and testing. A
**corridor** is a stretch of freight route and can contain several scenes; where it does, they are usually the
same ground photographed on different dates rather than different places. A scene is **labelled** if its vehicles
have been marked, which is true of 21 of the 24, and only labelled scenes contribute to any number here.
**Trained on** means that scene's vehicles were used to build that particular model. It differs between the two
models, so the same scene can be training data for one and unseen by the other.

---

_Source data: 24 PlanetScope SuperDove scenes, 789 labelled vehicles, two trained models and eight confidence
thresholds, in `data/results.csv` (336 rows) and `data/threshold_sweep.csv` (28 thresholds per model), both
derived from the per-detection records in `data/detections_*.csv`. Method described in the companion methodology
report._
