---
title: "Choosing the confidence threshold"
subtitle: "Where each model actually performs best — and why raising the threshold does not fix over-counting"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    number_sections: true
    theme: flatly
    df_print: kable
    fig_width: 10
    fig_height: 6
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dpi = 150, fig.retina = 2)
```

```{r libraries}
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(tidyr); library(tibble)
  library(ggplot2); library(forcats); library(scales)
})
MODELS <- c(MODEL_A = "adamiak-v2", MODEL_B = "warmup-v1")
PAL    <- c("adamiak-v2" = "#2a78d6", "warmup-v1" = "#eb6834")

theme_td <- function(base = 11) {
  theme_minimal(base_size = base) +
    theme(panel.grid.minor = element_blank(),
          plot.title    = element_text(face = "bold", size = base + 2),
          plot.subtitle = element_text(colour = "grey35", size = base - 1),
          plot.caption  = element_text(colour = "grey45", size = base - 2, hjust = 0),
          strip.text    = element_text(face = "bold"),
          legend.position = "top", legend.title = element_blank())
}
theme_set(theme_td())

sweep <- read_csv("data/threshold_sweep.csv", show_col_types = FALSE) |>
  mutate(model = recode(model, !!!MODELS), model = factor(model, levels = unname(MODELS)))
results <- read_csv("data/results.csv", show_col_types = FALSE) |>
  mutate(model = recode(model, !!!MODELS), model = factor(model, levels = unname(MODELS)))

peaks  <- sweep |> group_by(model) |> slice_max(mean_F1, n = 1) |> ungroup()
parity <- sweep |> group_by(model) |> slice_min(abs(count_ratio - 1), n = 1) |> ungroup()
```

# The question

The confidence threshold is a free parameter. It does not change what the model detects — only which detections
survive into the output — so it can be chosen after training, at no cost, and changed at any time.

Model B (`warmup-v1`) over-counts badly at the thresholds originally sampled: at 0.3 it returns roughly
**2.9 detections for every labelled vehicle**. The obvious remedy is to raise the threshold until the excess
disappears. The natural question is how far to go, and whether something near the top of the range — 0.9 or
0.95 — would both remove the over-counting and maximise accuracy.

This document answers that by sweeping the threshold from 0.30 to 0.975 in 0.025 steps for both models. No
re-inference was needed: every detection was stored with its score, so the sweep is a filter over records
already on disk.

# The short answer

**No — 0.95 is far too high for either model.** F1 peaks well below it and falls away steeply, and the
over-counting is replaced by severe under-counting rather than corrected.

```{r headline-answer}
sweep |>
  filter(threshold %in% c(0.30, 0.50, 0.65, 0.775, 0.85, 0.95)) |>
  select(model, threshold, mean_F1, pooled_F1, pooled_P, pooled_R, count_ratio) |>
  arrange(model, threshold) |>
  mutate(across(-c(model, threshold), \(x) round(x, 3)))
```

At 0.95, model B finds **fewer than a third** of the labelled trucks and model A finds about **one in twenty**.
Both are worse on F1 than they are at 0.3.

# The full sweep

```{r fig-sweep, fig.height=5.6, fig.width=11}
sweep |>
  select(model, threshold, `mean F1` = mean_F1, `count ratio` = count_ratio) |>
  pivot_longer(c(`mean F1`, `count ratio`), names_to = "metric", values_to = "value") |>
  ggplot(aes(threshold, value, colour = model)) +
  geom_hline(data = tibble(metric = "count ratio", y = 1), aes(yintercept = y),
             linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 1) +
  geom_point(data = peaks |> transmute(model, threshold, value = mean_F1, metric = "mean F1"),
             size = 3.4, shape = 21, stroke = 1.5, fill = "white") +
  geom_point(data = parity |> transmute(model, threshold, value = count_ratio, metric = "count ratio"),
             size = 3.4, shape = 21, stroke = 1.5, fill = "white") +
  facet_wrap(~ metric, scales = "free_y") +
  scale_colour_manual(values = PAL) +
  labs(title = "F1 and count ratio across the confidence range",
       subtitle = "Circles mark each model's F1 peak (left) and its count-parity point (right).",
       x = "confidence threshold", y = NULL,
       caption = "Dashed line on the right panel is count parity — detections equal to labelled vehicles.")
```

Both F1 curves are single-peaked. They rise as spurious detections are filtered out, turn over once real
detections start being filtered too, and then fall away sharply. The count ratio declines monotonically and
crosses parity at a different threshold from the F1 peak — the two objectives do not coincide.

# Where each model should run

```{r operating-points}
bind_rows(
  peaks  |> transmute(model, purpose = "maximise F1", threshold,
                      mean_F1 = round(mean_F1, 3), pooled_F1 = round(pooled_F1, 3),
                      pooled_P = round(pooled_P, 3), pooled_R = round(pooled_R, 3),
                      count_ratio = round(count_ratio, 2)),
  parity |> transmute(model, purpose = "match the count", threshold,
                      mean_F1 = round(mean_F1, 3), pooled_F1 = round(pooled_F1, 3),
                      pooled_P = round(pooled_P, 3), pooled_R = round(pooled_R, 3),
                      count_ratio = round(count_ratio, 2))
) |> arrange(model, desc(purpose))
```

**Neither model was being run at its best setting.** Model A peaks at **0.55** and model B at **0.65** — both
above the 0.5 the console defaults to, and well below the 0.85 that was sampled as the high end.

The two models reach count parity in very different places: model A at **0.45**, model B at **0.775**. That gap
is a direct measure of how much more freely B fires — it needs a far stricter cutoff to produce the same number
of detections.

## F1 optimum versus count parity

For model A the two objectives nearly coincide (0.55 and 0.45), so a single threshold serves both. For model B
they are far apart (0.65 and 0.775), and the choice depends on the deliverable:

```{r tradeoff-b}
sweep |>
  filter(model == "warmup-v1", threshold %in% c(0.65, 0.70, 0.75, 0.775, 0.80)) |>
  select(threshold, mean_F1, pooled_P, pooled_R, count_ratio) |>
  mutate(across(everything(), \(x) round(x, 3)))
```

Moving model B from its F1 optimum (0.65) to its parity point (0.775) costs only **0.015 mean F1** while taking
the count ratio from 1.49 to 1.03. For a deliverable that reports aggregate truck counts, that is a good trade:
a near-unbiased counter for almost no loss in detection quality.

# Why raising the threshold cannot fix over-counting

Raising the threshold removes detections indiscriminately — it cannot distinguish a spurious detection from a
real one, only a low-scoring one from a high-scoring one. Because real and spurious detections overlap heavily
in score, filtering away the spurious ones necessarily takes real ones with them.

```{r fig-pr-tradeoff, fig.height=5.4, fig.width=10}
sweep |>
  select(model, threshold, Precision = pooled_P, Recall = pooled_R, F1 = pooled_F1) |>
  pivot_longer(c(Precision, Recall, F1), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric, levels = c("Precision", "Recall", "F1"))) |>
  ggplot(aes(threshold, value, colour = model)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ metric, nrow = 1) +
  scale_colour_manual(values = PAL) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(title = "Precision rises, recall falls, F1 turns over",
       subtitle = "Pooled across all 21 labelled scenes.",
       x = "confidence threshold", y = NULL,
       caption = "Past each model's peak, recall is lost faster than precision is gained.")
```

Precision climbs steadily with the threshold and recall declines steadily. F1 combines them, so it rises only
while the precision gain outweighs the recall loss. Past the peak the balance inverts, and by 0.95 both models
have given up far more real detections than spurious ones.

Model A's collapse is the more extreme: its scores are concentrated lower, so a 0.95 cutoff removes almost
everything it produces.

# Consistency is not the same as accuracy

Spread keeps improving as the threshold rises, even where accuracy is collapsing.

```{r fig-sd, fig.height=5, fig.width=9}
sweep |>
  ggplot(aes(threshold, SD_F1, colour = model)) +
  geom_line(linewidth = 1) +
  geom_point(data = peaks, size = 3.2, shape = 21, stroke = 1.4, fill = "white") +
  scale_colour_manual(values = PAL) +
  labs(title = "Standard deviation of per-scene F1 across the confidence range",
       subtitle = "Circles mark each model's F1 peak. Lower SD is not automatically better.",
       x = "confidence threshold", y = "SD of per-scene F1",
       caption = "A model that detects almost nothing scores consistently badly — low spread, no value.")
```

This is worth stating explicitly because low variance was earlier identified as one of model B's strengths.
That reading holds where accuracy is reasonable, but variance alone is not a quality measure: at 0.95 model B's
F1 standard deviation is among its lowest while its F1 is near its worst.

# Effect on the model comparison

Re-running the head-to-head at each model's own optimum rather than a shared threshold narrows the gap:

```{r fig-at-optimum, fig.height=7, fig.width=10}
opt <- tibble(model = factor(c("adamiak-v2", "warmup-v1"), levels = unname(MODELS)),
              threshold = c(0.55, 0.65))

# nearest sampled thresholds available in results.csv
at_opt <- results |>
  filter((model == "adamiak-v2" & threshold == 0.5) | (model == "warmup-v1" & threshold == 0.65))

at_opt |>
  mutate(scene_id = fct_reorder(scene_id, annotated_count)) |>
  ggplot(aes(f1, scene_id, fill = model)) +
  geom_col(position = position_dodge(width = 0.76), width = 0.7) +
  scale_fill_manual(values = PAL) +
  labs(title = "Per-scene F1 with each model at its own best threshold",
       subtitle = "adamiak-v2 at 0.50, warmup-v1 at 0.65. Scenes ordered by annotation count.",
       x = "F1", y = NULL,
       caption = "Comparing models at a shared threshold penalises whichever one is further from its optimum.")
```

```{r optimum-summary}
at_opt |>
  group_by(model, threshold) |>
  summarise(mean_F1 = round(mean(f1), 3), SD_F1 = round(sd(f1), 3),
            mean_P = round(mean(precision), 3), mean_R = round(mean(recall), 3),
            median_count_ratio = round(median(count_ratio), 2), .groups = "drop")
```

Model A remains ahead on mean F1, but the margin is smaller than the shared-threshold comparison suggested, and
model B is closer to count parity at its own optimum than A is at the threshold the console currently uses.

# Recommendations

1. **Stop running either model at 0.5 by default.** It is below both optima and was inherited rather than chosen.
2. **For detection quality:** run model A at **0.55**, model B at **0.65**.
3. **For counting:** run model A at **0.45**, model B at **0.775**. Model B loses only 0.015 F1 moving to parity.
4. **Do not go above ~0.8 for either model.** Past that, real detections are lost faster than spurious ones and
   both accuracy and count fidelity degrade.
5. **Compare models at their own optima**, not a shared threshold, or the comparison measures threshold
   mismatch as much as model quality.

# Caveats

- **Precision is a lower bound.** Only clear, well-formed truck echoes were labelled, so some counted false
  positives are real unlabelled vehicles. True precision is higher than shown, which means the real F1 peak
  probably sits slightly *below* the thresholds identified here, and the true count ratios slightly lower.
- **These thresholds are fitted on the same 21 scenes they are evaluated on.** They are descriptive of this
  dataset, not validated operating points. A held-out corridor would be needed to confirm they transfer.
- **The parity points are dataset-wide.** Per-corridor parity differs substantially — see the over-detection
  breakdown in the main report — so a single global threshold still leaves per-corridor bias.
- **Matching is custom**: red-keypoint distance within 18 m, not box IoU, so these values are not comparable to
  published mAP figures.

Underlying data: `data/threshold_sweep.csv` (28 thresholds × 2 models) and `data/results.csv`
(21 scenes × 2 models × 6 thresholds), both derived from the stored per-detection records.
