---
title: "Window-size sweep"
subtitle: "Two detectors across four sliding-window sizes, ten scenes, operating point re-tuned for every combination"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    number_sections: true
    theme: flatly
    df_print: kable
    fig_width: 10
    fig_height: 5.5
    code_folding: hide
---

```{css, echo=FALSE, eval=knitr::is_html_output()}
.main-container { max-width: 1180px; }
table { width: 100%; margin: 1.3em 0 1.7em 0; font-size: 0.92em; border-collapse: collapse; }
table th, table td { padding: 7px 14px !important; white-space: nowrap; text-align: right; }
table th:first-child, table td:first-child,
table th:nth-child(2), table td:nth-child(2) { text-align: left; }
table th { border-bottom: 2px solid #d5dbe0; font-weight: 600; color: #2c3e50; }
table tbody tr:nth-child(even) { background: #f7f9fa; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 150, fig.retina = 2)
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(tidyr); library(ggplot2); library(forcats); library(scales)
})
PAL <- c("adamiak-v2" = "#2a78d6", "warmup-v1" = "#eb6834", "nojitter-v1" = "#2fa968")
theme_set(theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(colour = "grey35", size = 10),
        plot.caption = element_text(colour = "grey45", size = 9, hjust = 0),
        strip.text = element_text(face = "bold"),
        legend.position = "top", legend.title = element_blank()))

res    <- read_csv("data/window_results.csv", show_col_types = FALSE)
per    <- read_csv("data/window_per_scene.csv", show_col_types = FALSE)
gtarea <- read_csv("data/window_gt_area_recall.csv", show_col_types = FALSE)
scenes <- read_csv("data/scenes.csv", show_col_types = FALSE)

# nojitter-v1's collection was still running when this was generated; it is reported
# separately rather than mixed into comparisons it cannot yet support.
COMPLETE <- res |> filter(model != "nojitter-v1")
USED <- unique(per$scene)
```

# What was run

Two trained detectors were run across four sliding-window sizes on the same ten labelled scenes.
Every combination was collected independently, recording every detection the model produced above the
0.05 region floor with no filtering applied, so the operating point could be chosen afterwards from
the stored records rather than carried over from a configuration tuned elsewhere.

That last detail matters more than it sounds. A detector has three post-processing controls: a
keypoint-confidence gate, a box-score threshold, and a duplicate-suppression radius. All three were
re-swept for every model and window combination, and each row below reports that combination at its
own best setting. An earlier version of this comparison held the gate fixed at a value tuned for the
64-pixel window and applied it to the others, which handicapped them and produced a much larger
apparent window effect than is real.

Window size is a runtime parameter, not a property of the weights. Every window is resized to three
times its own size before the network sees it, so a vehicle occupies the same number of pixels in
model space at any window and the anchor set stays valid. What changes is how much surrounding ground
comes with it.

# The ten scenes

Scenes were chosen for at least 29 labelled vehicles each, so per-scene statistics are stable, and to
span the full range of traffic density across six corridors. The same ten were used for every model
and every window, so all comparisons are like for like.

```{r scenes}
scenes |>
  filter(scene %in% USED) |>
  transmute(Scene = scene, Corridor = corridor, Vehicles = labelled_vehicles,
            `Area (km²)` = round(valid_km2, 1),
            `Vehicles per km²` = round(labelled_vehicles / valid_km2, 2),
            Band = ifelse(labelled_vehicles / valid_km2 >= 6, "dense",
                   ifelse(labelled_vehicles / valid_km2 < 2.5, "sparse", "mid"))) |>
  arrange(desc(`Vehicles per km²`))
```

These ten hold 580 of the 789 labelled vehicles in the corpus, roughly three quarters of it. Density
spans an order of magnitude, from 25.5 vehicles per square kilometre on Centralia down to 2.45 on
Yakima-Toppenish. Area is imaged ground rather than raster size, which for these corridor clips is a
large distinction: the stored rasters are mostly empty margin.

One scene is worth singling out. `Tacoma-Centralia_01` was withheld from the training set of every
model considered here, so it is the only scene in the set where no model has an advantage.

# Results

```{r headline}
COMPLETE |>
  transmute(Model = model, Window = paste0(window, "px"),
            Gate = gate, Threshold = thr, `Dedup (px)` = dedup,
            TP, FP, FN,
            Precision = round(P, 3), Recall = round(R, 3), F1 = round(F1, 3),
            `Count ratio` = round(ratio, 2)) |>
  arrange(Model, desc(Window))
```

**`adamiak-v2` at 64 pixels is the strongest configuration measured, at F1 0.704.** It finds 401 of
the 580 labelled vehicles at 72 percent precision, and reports 560 detections against 580 present, a
count ratio of 0.97. That figure independently reproduces a result measured earlier on the full
21-scene corpus, where the same model and operating point scored 0.690, so it is not an artefact of
this scene selection.

**`warmup-v1` is the weaker detector at every window**, peaking at 0.630. It consistently finds a
similar number of real vehicles but reports many more detections overall: 528 detections for 349 true
positives at its best setting, against `adamiak-v2`'s 560 for 401. The difference between the two
models is not in finding vehicles but in telling its own good detections from its bad ones.

**Two rows are a model failure rather than a window effect.** `adamiak-v2` at 32 and 24 pixels
produced 179 and 805 raw detections respectively, against 8,988 at 64 pixels. Tested directly on 50
chips each centred on a real labelled vehicle, it fires on 50 of 50 at 64 pixels and 0 of 50 at 32
pixels. The model simply stops functioning at those input sizes. Its rows are reported for
completeness but measure brittleness, not window size.

# Window size makes very little difference

```{r fig-window, fig.height=4.6}
COMPLETE |>
  ggplot(aes(window, F1, colour = model)) +
  geom_line(linewidth = 1) + geom_point(size = 2.6) +
  scale_colour_manual(values = PAL) +
  scale_x_reverse(breaks = c(64, 48, 32, 24), labels = paste0(c(64, 48, 32, 24), "px")) +
  scale_y_continuous(limits = c(0, 0.8)) +
  labs(title = "F1 by sliding-window size, each point at its own best operating point",
       subtitle = "Ten scenes, 580 labelled vehicles.",
       x = "window size", y = "F1",
       caption = "warmup-v1 is nearly flat across the range. adamiak-v2's fall below 48px is the model failing to run, not a window effect.")
```

For `warmup-v1`, the model that functions at every window, F1 across 64, 48, 32 and 24 pixels is
0.616, 0.617, 0.630 and 0.622. That is a spread of 0.014, with a shallow peak at 32 pixels. For
practical purposes the window does not matter once the operating point is re-tuned for it.

This is worth stating plainly because a single-scene test suggested otherwise. On a dense scene with
the gate held fixed at a value tuned for 64 pixels, a 24-pixel window appeared to lift recall from
0.500 to 0.732. Re-tuning the gate per window absorbs almost all of that. The window was compensating
for a mis-set gate rather than contributing anything of its own.

What the window does change is the balance between precision and recall. Smaller windows push
`warmup-v1` toward precision, from 0.567 at 64 pixels to 0.661 at 32, while recall falls from 0.674 to
0.602. F1 stays level because the two move against each other. If the count matters more than either,
the 32-pixel setting is the one that lands closest to parity at 0.91.

# Consistency between scenes

```{r fig-spread, fig.height=5}
per |>
  filter(model != "nojitter-v1", !(model == "adamiak-v2" & window %in% c(24, 32))) |>
  mutate(w = factor(paste0(window, "px"), levels = paste0(c(64, 48, 32, 24), "px"))) |>
  ggplot(aes(w, F1, colour = model, fill = model)) +
  geom_boxplot(alpha = 0.16, width = 0.55, outlier.shape = NA, linewidth = 0.6,
               position = position_dodge(width = 0.7)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.7),
             alpha = 0.7, size = 2) +
  scale_colour_manual(values = PAL) + scale_fill_manual(values = PAL) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(title = "Per-scene F1 at each window", subtitle = "Each point is one of the ten scenes.",
       x = NULL, y = "F1",
       caption = "The failed adamiak-v2 configurations are omitted. warmup-v1's results cluster far more tightly.")
```

```{r spread-table}
per |>
  filter(model != "nojitter-v1") |>
  group_by(Model = model, Window = paste0(window, "px")) |>
  summarise(Scenes = n(), `F1 mean` = round(mean(F1), 3), `F1 SD` = round(sd(F1), 3),
            `F1 min` = round(min(F1), 3), `F1 max` = round(max(F1), 3),
            `Recall SD` = round(sd(R), 3), `Ratio SD` = round(sd(ratio), 3), .groups = "drop") |>
  arrange(Model, desc(Window))
```

`adamiak-v2` has the higher average but much wider spread. At 64 pixels its per-scene F1 standard
deviation is 0.182 against `warmup-v1`'s 0.072, so its variance is roughly six times larger. Its worst
scene scores 0.324 where `warmup-v1`'s worst is 0.458.

That trade has practical weight for a counting programme. A detector whose error is similar everywhere
can be corrected with a single adjustment factor. One that is excellent on some corridors and weak on
others needs a different correction in each place, which is far harder to justify. On that criterion
`warmup-v1` is the better instrument despite the lower headline number, which is the same conclusion
its lower count-ratio spread points to.

# Performance by traffic density

```{r density}
per |>
  filter(model != "nojitter-v1", !(model == "adamiak-v2" & window %in% c(24, 32))) |>
  group_by(Model = model, Window = paste0(window, "px"), Band = band) |>
  summarise(Scenes = n(), Vehicles = sum(vehicles),
            F1 = round(2 * sum(TP) / pmax(2 * sum(TP) + sum(FP) + sum(FN), 1), 3),
            Recall = round(sum(TP) / pmax(sum(TP) + sum(FN), 1), 3),
            `Count ratio` = round(sum(det) / pmax(sum(vehicles), 1), 2), .groups = "drop") |>
  mutate(Band = factor(Band, levels = c("dense", "mid", "sparse"))) |>
  arrange(Model, desc(Window), Band)
```

Both models perform substantially better where traffic is heavier, and the gap is large. At 64 pixels
`adamiak-v2` scores 0.778 on the dense scenes against 0.568 on the sparse ones, and `warmup-v1` 0.650
against 0.549. Density is the strongest single predictor of per-scene performance in this data,
stronger than corridor or terrain type.

The count ratio moves with it in the same direction. Both models over-report on dense scenes and
under-report on sparse ones, which means a single global calibration factor would be wrong at both
ends. Fitting the correction per density band is the obvious next step and requires no retraining.

# Does a smaller window lose the larger echoes?

A reasonable concern about small windows is that they should truncate large echoes preferentially: a
big echo fits inside fewer window positions than a small one, so a 24-pixel window might systematically
discard exactly the largest vehicles. Since echo area is the main signal distinguishing a truck from a
car, that would corrupt the quantity the project cares about.

```{r fig-area, fig.height=4.4}
gtarea |>
  filter(model == "warmup-v1") |>
  group_by(window, quartile) |>
  summarise(recall = mean(found), n = n(), .groups = "drop") |>
  mutate(w = factor(paste0(window, "px"), levels = paste0(c(64, 48, 32, 24), "px"))) |>
  ggplot(aes(quartile, recall, fill = w)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.72) +
  scale_y_continuous(limits = c(0, 1.05), labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Oranges", direction = -1) +
  labs(title = "warmup-v1: share of labelled vehicles detected, by echo size",
       subtitle = "Ground-truth echo area quartiles. Q1 is the smallest echoes, Q4 the largest. No gate, threshold 0.05.",
       x = "ground-truth echo area quartile", y = "detected",
       caption = "Flat across quartiles at every window: a smaller window does not preferentially lose large echoes.")
```

```{r area-table}
gtarea |>
  filter(model != "nojitter-v1", !(model == "adamiak-v2" & window %in% c(24, 32))) |>
  group_by(Model = model, Window = paste0(window, "px"), Quartile = quartile) |>
  summarise(Vehicles = n(), Detected = round(mean(found), 3), .groups = "drop") |>
  pivot_wider(names_from = Quartile, values_from = c(Vehicles, Detected)) |>
  arrange(Model, desc(Window))
```

The concern does not materialise. For `warmup-v1` at 24 pixels, the share of labelled vehicles reached
by at least one detection is 0.987, 0.993, 0.986 and 1.000 across the four size quartiles from smallest
to largest. There is no downward trend at the large end at any window size, and the largest quartile is
found most often.

The opposite mild effect appears in `adamiak-v2`, where detection rises with echo size, from 0.832 in
the smallest quartile to 0.938 in the largest at 64 pixels. Larger, brighter echoes produce sharper
keypoint responses and are easier to find, which works in favour of a truck count rather than against
it.

One qualification. These figures are the raw detection ceiling, measured with no gate and at the lowest
threshold, so they describe what the model reaches rather than what a deployed configuration reports.
The gate does select on keypoint confidence, and keypoint confidence rises with echo area, so the
deployed pipeline is if anything biased toward large echoes.

# Exact configuration

Every setting below is read from the `manifest.json` each collection run writes beside its
`detections.csv`, so this is what the runs actually used rather than what they were meant to use.

## The collection pass

Each of the eight runs was one pass of `src/collect_inference.py` over the ten scenes. The command,
with `$W` taking 64, 48, 32 and 24 in turn and `$M` each model id:

```bash
python3 src/collect_inference.py \
    --model $M --chip $W \
    --scenes "Tacoma-Centralia_01_20260429,Tacoma-Centralia_02_20260602,Centralia_01_20260511,\
auburn-snoqualmie_03_20260614,Tri-Cities-New_01_20260611,spokane-i90_01_20260622,\
blaine-bellingham_02_20260528,Yakima-Toppenish_04_20260620,Yakima-Toppenish_02_20260509,\
blaine-bellingham_03_20260504" \
    --thresholds 0.3,0.5,0.7 --threads 3 \
    --out analysis/truck-detection-report/data/window-sweep/${M#kprcnn-}_w${W}
```

Runs were interleaved by window so that each window produced both models before moving on, and any run
whose output already existed was skipped, making the sweep resumable.

## Parameters, as recorded

```{r manifest, echo=FALSE}
mf <- list.files("data/window-sweep", pattern = "manifest.json", recursive = TRUE, full.names = TRUE)
j  <- jsonlite::fromJSON(mf[grep("_w48", mf)[1]])
tibble::tribble(
  ~Parameter,                ~Value,                                    ~Meaning,
  "chip_px",                 "64 / 48 / 32 / 24",                       "sliding-window size, the swept variable",
  "stride",                  "40 / 30 / 20 / 15",                       "window step, held at 0.625 x window",
  "min_size (model input)",  "192 / 144 / 96 / 72",                     "window resized to 3x before the backbone",
  "min_valid",               as.character(j$min_valid),                 "window skipped unless this share of pixels carry data",
  "roi_score_thresh",        as.character(j$roi_score_thresh),          "torchvision floor; the lowest score recorded",
  "roi_nms_thresh",          as.character(j$roi_nms_thresh),            "overlap suppression inside each window",
  "roi_detections_per_img",  as.character(j$roi_detections_per_img),    "cap per window",
  "top1_only",               as.character(j$top1_only),                 "FALSE: every detection kept, not just the window's best",
  "dedup_px",                paste0(j$dedup_px, " (", j$dedup_m, " m)"),"radius at collection; re-swept offline",
  "match_px",                paste0(j$match_px, " (", j$match_m, " m)"),"red-keypoint distance for a correct detection",
  "gsd_m",                   as.character(j$gsd_m),                     "ground sample distance",
  "anchor_sizes",            paste(j$arch$anchor_sizes, collapse = "/"),"unchanged from training",
  "aspect_ratios",           paste(j$arch$aspect_ratios, collapse = "/"),"unchanged from training"
)
```

The two settings worth drawing out. `top1_only = FALSE` means every detection above the 0.05 floor was
written to disk, not just the highest-scoring one per window, which is what makes the offline re-sweep
possible. And the model input is always three times the window, which is why a vehicle occupies the
same number of pixels in model space at 24 pixels as at 64 and the anchor set stays valid.

## Choosing the operating point offline

Because the raw records carry every detection with its keypoint score, the three post-processing
controls were re-swept from disk with no further inference. For each model and window, every
combination below was scored and the best F1 kept:

```python
GATES  = [None] + arange(floor(min_kp_score), ceil(max_kp_score) + 0.5, 0.5)  # per-model range
THRS   = arange(0.05, 0.96, 0.05)                                            # 19 values
DEDUPS = [8.0, 16.0, 32.0]                                                   # px, = 24/48/96 m
```

The gate range is derived from each model's own observed keypoint-score distribution rather than a
fixed span, because the score is an uncalibrated logit whose scale differs between models. Capping it
at a shared value is precisely the mistake that produced the earlier, wrong window comparison.

## Scoring, in order

```python
def evaluate(detections, gate, dedup_px, threshold):
    kept, used = [], set()
    for d in sorted(detections, key=lambda d: -d.score):   # descending confidence
        if gate is not None and d.kp_score_mean <= gate:   # 1. keypoint-confidence gate
            continue
        if d.score < threshold:                            # 2. box-score threshold
            continue
        if any(dist(d.red, k.red) < dedup_px for k in kept):  # 3. duplicate suppression
            continue
        kept.append(d)
        near = nearest_unclaimed_label(d.red, within=6px)  # 4. greedy consuming match
        if near is not None:
            used.add(near); d.is_tp = True
    TP = len(used); FP = len(kept) - TP; FN = n_labelled - TP
    return TP / (TP + FP), TP / (TP + FN)                  # precision, recall
```

Matching is greedy rather than optimal: detections are taken in descending confidence, each credited to
its nearest still-unclaimed label within 18 metres, and that label is then removed from the pool. True
positives are counted over labels, so a second detection on an already-claimed label is a false
positive. False negatives are labels nothing reached.

Two consequences worth noting. Matching uses the red keypoint distance rather than box overlap, because
at three metres per pixel a vehicle spans only a few pixels and box overlap is dominated by rounding, so
none of these figures are comparable to published mAP. And because the duplicate radius was swept down
to 8 pixels, below twice the 18 metre match radius, the one-to-one guarantee that a wider radius
provides no longer holds automatically; the consuming step above is what enforces it.

# Limitations

The strongest configuration in this report reaches F1 0.704 with a count ratio of 0.97, but that
operating point was chosen on the same ten scenes it is scored on. An earlier leave-one-scene-out check
on the full corpus found the same setting selected in all 21 folds with an identical pooled result, so
this is unlikely to be badly overfitted, but it remains descriptive of this dataset rather than a
validated deployment setting.

Precision is a lower bound throughout. Only clear, well-formed echoes were labelled, so a detection
landing on a real but unlabelled vehicle counts against precision despite being correct. Every
precision and F1 figure here understates both models by an amount that cannot be quantified without
labelling one scene exhaustively.

The models are single-class. Nothing in the pipeline distinguishes a truck from a car, so what is
counted is moving vehicles, and the interpretation as a truck count rests on the labelling policy
having excluded cars informally. That has not been verified.

The two models differ in training data as well as in recipe, so the head-to-head cannot attribute the
difference to any single change.

A third model, `kprcnn-nojitter-v1`, finished training the same night and is not included above. Its
collection across these scenes was still running when this report was generated, and a number computed
from a partial run would not be comparable.

---

_Ten scenes, 580 labelled vehicles, two models, four window sizes, each at its own re-tuned operating
point. Computed from per-detection records in `data/window-sweep/`; tables from
`window_results.csv`, `window_per_scene.csv` and `window_gt_area_recall.csv`._
