---
title: "Window-size sweep"
subtitle: "Two detectors, four sliding-window sizes, ten scenes, operating point re-tuned for every combination"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    number_sections: true
    theme: flatly
    df_print: kable
    fig_width: 10
    fig_height: 5
    code_folding: hide
---

```{css, echo=FALSE, eval=knitr::is_html_output()}
.main-container { max-width: 1180px; }
table { width: 100%; margin: 1.2em 0 1.6em 0; font-size: 0.92em; border-collapse: collapse; }
table th, table td { padding: 7px 14px !important; white-space: nowrap; text-align: right; }
table th:first-child, table td:first-child,
table th:nth-child(2), table td:nth-child(2) { text-align: left; }
table th { border-bottom: 2px solid #d5dbe0; font-weight: 600; color: #2c3e50; }
table tbody tr:nth-child(even) { background: #f7f9fa; }
.lede { font-size: 1.05em; line-height: 1.62; border-left: 3px solid #dda032;
        padding: 4px 0 4px 16px; margin: 16px 0 28px; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 150, fig.retina = 2)
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(tidyr); library(ggplot2); library(forcats); library(scales)
})
PAL <- c("adamiak-v2" = "#2a78d6", "warmup-v1" = "#eb6834", "nojitter-v1" = "#2fa968")
theme_set(theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(colour = "grey35", size = 10),
        plot.caption = element_text(colour = "grey45", size = 9, hjust = 0),
        strip.text = element_text(face = "bold"),
        legend.position = "top", legend.title = element_blank()))

res    <- read_csv("data/window_results.csv", show_col_types = FALSE)
per    <- read_csv("data/window_per_scene.csv", show_col_types = FALSE)
gtarea <- read_csv("data/window_gt_area_recall.csv", show_col_types = FALSE)
scenes <- read_csv("data/scenes.csv", show_col_types = FALSE)
rb     <- read_csv("data/window_robustness.csv", show_col_types = FALSE)

# nojitter-v1's collection was still running when this was generated, so it is left out
# rather than compared on partial data.
COMPLETE <- res |> filter(model != "nojitter-v1")
USED <- unique(per$scene)
# Configurations where the model does not actually run are kept everywhere but flagged,
# so nothing is dropped without saying so.
per <- per |> mutate(works = !(model == "adamiak-v2" & window %in% c(24, 32)))
```

::: lede
**`adamiak-v2` at 64 px is the best configuration measured: F1 0.704 at a count ratio of 0.97.** Window
size makes almost no difference once the operating point is re-tuned for it, a spread of 0.014 across
64 down to 24 px. What matters more is that **only three of eleven models can change window at all**
without failing outright. `adamiak-v2` scores higher than `warmup-v1` at every working window, but
`warmup-v1`'s per-scene variance is two and a half times tighter, which for a counting programme is the
more useful property. Traffic density predicts performance more strongly than either.
:::

# Results

```{r headline}
COMPLETE |>
  transmute(Model = model, Window = paste0(window, "px"),
            Gate = gate, Threshold = thr, `Dedup (px)` = dedup, TP, FP, FN,
            Precision = round(P, 3), Recall = round(R, 3), F1 = round(F1, 3),
            `Count ratio` = round(ratio, 2)) |>
  arrange(Model, desc(Window))
```

Every row is that model and window at its own best operating point, found by re-sweeping the keypoint
gate, the score threshold and the duplicate radius over the stored detection records. Ten scenes, 580
labelled vehicles.

`adamiak-v2` at 64 px finds 401 of the 580 vehicles at 72 percent precision, reporting 560 detections
against 580 present. That reproduces a figure measured separately on the full 21-scene corpus, where
the same model and operating point scored 0.690, so it is not an artefact of this scene selection.

`warmup-v1` peaks at 0.630. It finds a similar number of real vehicles but reports many more detections
overall, 528 for 349 true positives against `adamiak-v2`'s 560 for 401. The gap between them is not in
finding vehicles but in separating good detections from bad ones.

The `adamiak-v2` rows at 32 and 24 px are a model failure rather than a window effect, explained in
section 3.

# Window size makes almost no difference

```{r fig-window, fig.height=4.4, fig.width=11}
# Window is an ordered factor, not a continuous axis: the four sizes are the levels tested,
# not samples of a continuum, and even spacing stops 32/24 from crowding.
# Precision, recall and F1 share one 0-1 axis because they are the same unit; a free axis
# per panel would turn a 0.06 range into a dramatic slope.
wf <- function(d) factor(paste0(d$window, "px"), levels = paste0(c(64, 48, 32, 24), "px"))
pr <- COMPLETE |>
  mutate(w = wf(COMPLETE), works = !(model == "adamiak-v2" & window %in% c(24, 32))) |>
  select(model, w, works, Precision = P, Recall = R, F1) |>
  pivot_longer(c(Precision, Recall, F1), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric, levels = c("Precision", "Recall", "F1")))

ggplot(pr, aes(w, value, colour = model, group = model)) +
  geom_line(linewidth = 0.9, alpha = 0.45, linetype = "22") +
  geom_line(data = filter(pr, works), linewidth = 1) +
  geom_point(aes(shape = works), size = 2.6, fill = "white", stroke = 1.1) +
  facet_wrap(~ metric, nrow = 1) +
  scale_shape_manual(values = c(`TRUE` = 21, `FALSE` = 4), guide = "none") +
  scale_colour_manual(values = PAL) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2), labels = percent_format(accuracy = 1)) +
  labs(title = "Precision, recall and F1 by window size",
       subtitle = "Common 0-100% axis across all three panels. Each point at its own best operating point.",
       x = NULL, y = NULL,
       caption = "Crosses on a dashed segment are configurations where the model does not run; they are not a trend.")
```

```{r fig-ratio, fig.height=3.6, fig.width=7}
COMPLETE |>
  mutate(w = wf(COMPLETE), works = !(model == "adamiak-v2" & window %in% c(24, 32))) |>
  ggplot(aes(w, ratio, colour = model, group = model)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.9, alpha = 0.45, linetype = "22") +
  geom_line(data = ~ filter(.x, works), linewidth = 1) +
  geom_point(aes(shape = works), size = 2.6, fill = "white", stroke = 1.1) +
  scale_shape_manual(values = c(`TRUE` = 21, `FALSE` = 4), guide = "none") +
  scale_colour_manual(values = PAL) +
  scale_y_continuous(limits = c(0, 1.4), breaks = seq(0, 1.4, 0.2)) +
  annotate("text", x = 0.6, y = 1, label = "parity", vjust = -0.5, hjust = 0,
           size = 3.1, colour = "grey40") +
  labs(title = "Detections reported per labelled vehicle",
       subtitle = "Shown separately: a ratio, not a rate, so it does not share the axis above.",
       x = NULL, y = "detections ÷ vehicles present",
       caption = "warmup-v1 crosses parity between 48 and 32 px. adamiak-v2 sits just under it at its working windows.")
```

For `warmup-v1`, the model that functions at every window, F1 across 64, 48, 32 and 24 px is 0.616,
0.617, 0.630 and 0.622. A spread of 0.014, with a shallow peak at 32.

The first two panels show why. As the window shrinks, precision climbs from 0.567 to 0.661 and recall
falls from 0.674 to 0.602 by almost the same amount. Nothing is gained or lost overall; the operating
point slides along the same precision-recall curve. What does move usefully is the count, which crosses
parity between 48 and 32 px to land at 0.91, against 1.19 at 64.

This is worth stating because a single-scene test suggested the opposite. On a dense scene with the
gate held at a value tuned for 64 px, a 24 px window appeared to lift recall from 0.500 to 0.732.
Re-tuning the gate per window absorbs nearly all of that. The window was compensating for a mis-set
gate rather than contributing anything itself.

# Most models cannot change window at all

Before treating window size as a tunable parameter it is worth checking which models tolerate it. Every
registered model was given 60 chips, each centred on a real labelled vehicle, at all four windows. The
measure is whether it returns any detection above 0.3 on a chip that definitely contains a vehicle.

```{r robust, fig.height=4.8}
rb |>
  mutate(model = fct_reorder(model, hits, .fun = sum),
         w = factor(paste0(window, "px"), levels = paste0(c(64, 48, 32, 24), "px"))) |>
  ggplot(aes(w, model, fill = hits / n)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = hits, colour = hits > 35), size = 3.2, show.legend = FALSE) +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
  scale_fill_gradient(low = "#f7f2ea", high = "#2a78d6", labels = percent_format(),
                      limits = c(0, 1), name = NULL) +
  labs(title = "Detections on 60 chips centred on real vehicles",
       subtitle = "A model that cannot fire on a centred vehicle cannot be evaluated at that window.",
       x = "window size", y = NULL,
       caption = "Only jitter-mv, warmup-v1 and nojitter-v1 hold up across the full range.")
```

Three models hold up across the range: `jitter-mv`, `warmup-v1` and `nojitter-v1`. Every other model
degrades or fails outright below 48 px. `adamiak-v2` goes from 58 of 60 at its native window to 0 of 60
at 32 px. `adamiak-all` collapses a step earlier, 53 to 6 at 48 px.

The three that survive are not the three largest training sets, which rules out data volume as the
explanation: `jitter-mv` holds up on 411 chips while `adamiak-all` fails on 538. What they share is
that all three were trained from the **padded 96 px export**, the one carrying multi-vehicle targets.
In those chips a neighbouring vehicle can sit anywhere in the frame, so the model sees off-centre
vehicles during training even when translation jitter is switched off. That appears to be what lets a
model tolerate a change of input size.

The practical consequence is that window size is a parameter only for those three. For the rest it is a
fixed property of how they were built, and evaluating them at 32 or 24 px measures a failure mode
rather than a preference.

# Consistency between scenes

```{r box-helper, include=FALSE}
box_metric <- function(var, title, cap) {
  d <- per |>
    filter(model != "nojitter-v1") |>
    mutate(w = factor(paste0(window, "px"), levels = paste0(c(64, 48, 32, 24), "px")))
  # Dodge on `model` ONLY. Mapping linetype or shape adds them to the dodge grouping,
  # which splits each model into two slots and pulls its box away from its own points.
  # Failed configurations are marked with an annotation layer instead, outside the
  # positional aesthetics, at the dodged x offset (2 models, dodge width 0.7 -> +/-0.175).
  fail <- d |> filter(!works) |> distinct(w) |>
    mutate(x = as.numeric(w) - 0.175, y = 0.035)
  ggplot(d, aes(w, .data[[var]], colour = model, fill = model)) +
    geom_boxplot(alpha = 0.16, width = 0.55, outlier.shape = NA, linewidth = 0.6,
                 position = position_dodge(width = 0.7)) +
    geom_point(position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.7),
               alpha = 0.75, size = 1.9) +
    geom_text(data = fail, aes(x = x, y = y, label = "model does not run"),
              inherit.aes = FALSE, size = 2.7, colour = "#2a78d6", fontface = "italic") +
    scale_colour_manual(values = PAL) + scale_fill_manual(values = PAL) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25),
                       labels = percent_format(accuracy = 1)) +
    labs(title = title, subtitle = "Each point is one of the ten scenes; boxes show the middle half.",
         x = NULL, y = NULL, caption = cap)
}
```

```{r fig-box-f1, fig.height=4.3}
box_metric("F1", "Per-scene F1",
  "warmup-v1 clusters far more tightly at every window. The two adamiak-v2 configurations that do not run are labelled; their spread is a failure to fire, not a consistency measurement.")
```

```{r spread-table}
per |>
  filter(model != "nojitter-v1") |>
  mutate(Model = ifelse(works, model, paste0(model, " *"))) |>
  group_by(Model, Window = paste0(window, "px")) |>
  summarise(Scenes = n(), `F1 mean` = round(mean(F1), 3), `F1 SD` = round(sd(F1), 3),
            `F1 min` = round(min(F1), 3), `F1 max` = round(max(F1), 3),
            `Recall SD` = round(sd(R), 3), `Ratio SD` = round(sd(ratio), 3), .groups = "drop") |>
  arrange(Model, desc(Window))
```

`adamiak-v2` has the higher average and much wider spread. At 64 px its per-scene F1 standard deviation
is 0.182 against `warmup-v1`'s 0.072, so its variance is roughly six times larger, and its worst scene
scores 0.324 where `warmup-v1`'s worst is 0.458.

That trade has practical weight. A detector whose error is similar everywhere can be corrected with a
single adjustment factor. One that is excellent on some corridors and weak on others needs a different
correction in each place. On that criterion `warmup-v1` is the better instrument despite the lower
headline number, and its tighter count-ratio spread points the same way.

```{r fig-box-prec, fig.height=4.3}
box_metric("P", "Per-scene precision",
  "adamiak-v2 is the more precise model at both working windows, and its spread is wider at both ends.")
```

```{r fig-box-recall, fig.height=4.3}
box_metric("R", "Per-scene recall",
  "Recall is where both models are least consistent. The low outliers are the sparse scenes.")
```

Recall is where both are least consistent, and the low outliers are the sparse scenes, where a handful
of vehicles across a large area means a few misses move the number a long way. That spread is the
largest single contributor to the F1 variance above.

# Traffic density predicts performance

```{r density}
per |>
  filter(model != "nojitter-v1") |>
  mutate(Model = ifelse(works, model, paste0(model, " *"))) |>
  group_by(Model, Window = paste0(window, "px"), Band = band) |>
  summarise(Scenes = n(), Vehicles = sum(vehicles),
            F1 = round(2 * sum(TP) / pmax(2 * sum(TP) + sum(FP) + sum(FN), 1), 3),
            Recall = round(sum(TP) / pmax(sum(TP) + sum(FN), 1), 3),
            `Count ratio` = round(sum(det) / pmax(sum(vehicles), 1), 2), .groups = "drop") |>
  mutate(Band = factor(Band, levels = c("dense", "mid", "sparse"))) |>
  arrange(Model, desc(Window), Band)
```

At 64 px `adamiak-v2` scores 0.778 on the dense scenes against 0.568 on the sparse ones, and
`warmup-v1` 0.650 against 0.549. Density is the strongest single predictor of per-scene performance
here, stronger than corridor or terrain.

The count ratio moves the same way. Both models over-report where traffic is heavy and under-report
where it is sparse, so a single global calibration factor would be wrong at both ends. Fitting the
correction per density band is the obvious next step and needs no retraining.

Rows marked `*` are the two `adamiak-v2` configurations where the model does not run; their numbers are
a failure to fire rather than a property of the band.

# A smaller window does not lose the larger echoes

A reasonable objection to small windows is that they should truncate large echoes preferentially, since
a big echo fits inside fewer window positions than a small one. Because echo area is the main signal
separating a truck from a car, that would corrupt the quantity the project cares about.

```{r fig-area, fig.height=4.2}
gtarea |>
  filter(model == "warmup-v1") |>
  group_by(window, quartile) |>
  summarise(recall = mean(found), .groups = "drop") |>
  mutate(w = factor(paste0(window, "px"), levels = paste0(c(64, 48, 32, 24), "px"))) |>
  ggplot(aes(quartile, recall, fill = w)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.72) +
  scale_y_continuous(limits = c(0, 1.05), labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Oranges", direction = -1) +
  labs(title = "warmup-v1: share of labelled vehicles detected, by echo size",
       subtitle = "Ground-truth echo area quartiles, Q1 smallest to Q4 largest. No gate, threshold 0.05.",
       x = "ground-truth echo area quartile", y = "detected",
       caption = "Flat across quartiles at every window: a smaller window does not preferentially lose large echoes.")
```

It does not happen. For `warmup-v1` at 24 px the share of labelled vehicles reached by at least one
detection is 0.987, 0.993, 0.986 and 1.000 from smallest quartile to largest. No downward trend at the
large end at any window, and the largest quartile is found most often.

The opposite mild effect appears in `adamiak-v2`, where detection rises with echo size from 0.832 to
0.938 at 64 px. Larger, brighter echoes produce sharper keypoint responses and are easier to find,
which favours a truck count rather than working against it. Since the gate selects on keypoint
confidence and that rises with area, the deployed pipeline is if anything biased toward large echoes.

These are raw detection ceilings, measured with no gate at the lowest threshold, so they describe what
the model reaches rather than what a deployed configuration reports.

# The ten scenes

Chosen for at least 29 labelled vehicles each, so per-scene statistics are stable, and to span the
density range across six corridors. The same ten were used for every model and window.

```{r scenes}
scenes |>
  filter(scene %in% USED) |>
  transmute(Scene = scene, Corridor = corridor, Vehicles = labelled_vehicles,
            `Area (km²)` = round(valid_km2, 1),
            `Vehicles per km²` = round(labelled_vehicles / valid_km2, 2),
            Band = ifelse(labelled_vehicles / valid_km2 >= 6, "dense",
                   ifelse(labelled_vehicles / valid_km2 < 2.5, "sparse", "mid"))) |>
  arrange(desc(`Vehicles per km²`))
```

These hold 580 of the 789 labelled vehicles in the corpus. Density spans an order of magnitude, 25.5
vehicles per square kilometre down to 2.45. Area is imaged ground rather than raster size, which for
these corridor clips is a large distinction: the stored rasters are mostly empty margin.
`Tacoma-Centralia_01` was withheld from every model's training set, so it is the only scene here where
no model has an advantage.

# Exact configuration

Values are read from the `manifest.json` each collection run writes beside its `detections.csv`, so
this is what the runs used rather than what they were meant to use.

```bash
python3 src/collect_inference.py \
    --model $M --chip $W \
    --scenes "Tacoma-Centralia_01_20260429,Tacoma-Centralia_02_20260602,Centralia_01_20260511,\
auburn-snoqualmie_03_20260614,Tri-Cities-New_01_20260611,spokane-i90_01_20260622,\
blaine-bellingham_02_20260528,Yakima-Toppenish_04_20260620,Yakima-Toppenish_02_20260509,\
blaine-bellingham_03_20260504" \
    --thresholds 0.3,0.5,0.7 --threads 3 \
    --out analysis/truck-detection-report/data/window-sweep/${M#kprcnn-}_w${W}
```

```{r manifest, echo=FALSE}
mf <- list.files("data/window-sweep", pattern = "manifest.json", recursive = TRUE, full.names = TRUE)
j  <- jsonlite::fromJSON(mf[grep("_w48", mf)[1]])
tibble::tribble(
  ~Parameter,                ~Value,                                    ~Meaning,
  "chip_px",                 "64 / 48 / 32 / 24",                       "sliding-window size, the swept variable",
  "stride",                  "40 / 30 / 20 / 15",                       "window step, held at 0.625 x window",
  "min_size (model input)",  "192 / 144 / 96 / 72",                     "window resized to 3x before the backbone",
  "min_valid",               as.character(j$min_valid),                 "window skipped unless this share of pixels carry data",
  "roi_score_thresh",        as.character(j$roi_score_thresh),          "torchvision floor; the lowest score recorded",
  "roi_nms_thresh",          as.character(j$roi_nms_thresh),            "overlap suppression inside each window",
  "top1_only",               as.character(j$top1_only),                 "FALSE: every detection kept, not just the window's best",
  "dedup_px",                paste0(j$dedup_px, " (", j$dedup_m, " m)"),"radius at collection; re-swept offline",
  "match_px",                paste0(j$match_px, " (", j$match_m, " m)"),"red-keypoint distance for a correct detection",
  "anchor_sizes",            paste(j$arch$anchor_sizes, collapse = "/"),"unchanged from training"
)
```

Two settings carry the method. `top1_only = FALSE` writes every detection above the 0.05 floor to disk,
not just each window's best, which is what makes the offline re-sweep possible. And the model input is
always three times the window, so a vehicle occupies the same pixels in model space at 24 px as at 64
and the anchor set stays valid.

```python
GATES  = [None] + arange(floor(min_kp_score), ceil(max_kp_score) + 0.5, 0.5)  # per-model range
THRS   = arange(0.05, 0.96, 0.05)
DEDUPS = [8.0, 16.0, 32.0]        # px, = 24/48/96 m

def evaluate(detections, gate, dedup_px, threshold):
    kept, used = [], set()
    for d in sorted(detections, key=lambda d: -d.score):             # descending confidence
        if gate is not None and d.kp_score_mean <= gate: continue    # 1. keypoint gate
        if d.score < threshold: continue                             # 2. score threshold
        if any(dist(d.red, k.red) < dedup_px for k in kept): continue  # 3. dedup
        kept.append(d)
        near = nearest_unclaimed_label(d.red, within=6px)             # 4. consuming match
        if near is not None: used.add(near)
    TP = len(used); FP = len(kept) - TP; FN = n_labelled - TP
```

The gate range is derived per model from its own keypoint-score distribution rather than a fixed span,
because the score is an uncalibrated logit whose scale differs between models. Capping it at a shared
value is what produced an earlier, incorrect window comparison.

Matching is greedy, not optimal, and counts true positives over labels, so a second detection on an
already-claimed label is a false positive. It uses red-keypoint distance rather than box overlap,
because at three metres per pixel a vehicle spans a few pixels and box overlap is dominated by
rounding. None of these figures are comparable to published mAP.

# Limitations

The best configuration reaches F1 0.704 at a count ratio of 0.97, but that operating point was chosen
on the same ten scenes it is scored on. A leave-one-scene-out check on the full corpus selected the
same setting in all 21 folds with an identical pooled result, so it is unlikely to be badly overfitted,
but it remains descriptive of this dataset rather than a validated deployment setting.

Precision is a lower bound throughout. Only clear, well-formed echoes were labelled, so a detection
landing on a real but unlabelled vehicle counts against precision despite being correct.

The models are single-class. Nothing distinguishes a truck from a car, so what is counted is moving
vehicles, and reading it as a truck count rests on the labelling policy having excluded cars
informally. That has not been verified.

The two models differ in training data as well as recipe, so the head-to-head cannot attribute the
difference to any single change.

`kprcnn-nojitter-v1` finished training the same night and is not included; its collection across these
scenes was still running when this was generated.

---

_Ten scenes, 580 labelled vehicles, two models, four window sizes, each at its own re-tuned operating
point. Computed from per-detection records in `data/window-sweep/`._
